---
title: "Lecture 2: Fitting the Simple Linear Model"
subtitle: "PSTAT 126: Linear Regression"
author: "Annie Adams"
institute: "UC Santa Barbara"
date: "9/30/2026"
format:
  revealjs:
    theme: [default, ../styles.scss]
    slide-number: true
    preview-links: auto
    footer: "PSTAT 126 — Fall 2026 · UC Santa Barbara"
    logo: ""
    title-slide-attributes:
      data-background-color: "#003660"
---

## Review: The Simple Linear Model

From Monday, the simple linear regression model is:

$$y_i = \beta_0 + \beta_1 x_i + \varepsilon_i, \quad \varepsilon_i \overset{\text{iid}}{\sim} \mathcal{N}(0, \sigma^2)$$

- $\beta_1$ is a **comparison** between units 
- The fitted line $\widehat{y} = \widehat{\beta}_0 + \widehat{\beta}_1 x$ summarizes the relationship

**Today:** How do we estimate $\widehat{\beta}_0$, $\widehat{\beta}_1$, and $\sigma$ from data?

---

## Learning Objectives

By the end of today you will be able to:

- Explain the least squares criterion and apply it conceptually
- State the OLS formulas for the slope and intercept
- Compute and interpret fitted values and residuals
- Interpret the residual standard deviation $\widehat{\sigma}$ and $R^2$
- Explain what a standard error measures
- Use `lm()` in R to fit a simple linear regression and read its output

---

## Least Squares Estimation {background-color="#003660"}



## Mini Refresher: Derivatives {.smaller}

**Power rule:** $\frac{d}{dx} x^n = nx^{n-1}$

**Chain rule:** $\frac{d}{dx} f(g(x)) = f'(g(x)) \cdot g'(x)$ 

**Partial derivatives:** treat all other variables as constants and differentiate with respect to one variable at a time.

**Example:** $\dfrac{\partial}{\partial \beta_0}(y - \beta_0 - \beta_1 x)^2 = 2(y - \beta_0 - \beta_1 x) \cdot (-1)$


## Practice Problems

:::{.fragment}
::: callout-tip
## Practice Problems

:::{.nonincremental}
::: {style="font-size: 30px"}

Find the derivates of the following:

1.$x^6$

2. $y = (x^2 + 1)^3$

Find the partial derivative of the following:

1. $f(\beta_0, \beta_1) = (y - \beta_0 - \beta_1 x)^2$, find $\dfrac{\partial f}{\partial \beta_1}$

```{r}
#| echo: false
countdown::countdown(
  minutes = 2,
  play_sound = TRUE,
  color_border              = "#FEBC11",
  color_text                = "#003660",
  color_running_background  = "#003660",
  color_running_text        = "#FFFFFF",
  color_finished_background = "#FEBC11",
  color_finished_text       = "#003660",
  font_size = "2em",
)
```

:::
:::
:::
:::


---

## Mini Refresher: Summation Rules {.smaller}

**Derivatives through sums:** 
$$\frac{\partial}{\partial \beta_0}\sum_{i=1}^n f_i(\beta_0) = \sum_{i=1}^n \frac{\partial}{\partial \beta_0} f_i(\beta_0)$$

**Constants in sums:**

- $\displaystyle\sum_{i=1}^n c = nc$ — a constant summed $n$ times is just $n \times c$
- $\displaystyle\sum_{i=1}^n c \cdot x_i = c\sum_{i=1}^n x_i$ — a constant factor pulls out of the sum

::: {.callout-note}
These are all we need to minimize RSS — apply the chain rule, pass the derivative through the sum, set to zero, and simplify.
:::


## Practice Problems
:::{.fragment}
::: callout-tip
## Practice Problems

:::{.nonincremental}
::: {style="font-size: 30px"}

Simplify the following:

1. $\displaystyle\sum_{i=1}^{5} 4$

2. $\displaystyle\sum_{i=1}^{n} 3x_i$

3. $\dfrac{\partial}{\partial \beta_0}\displaystyle\sum_{i=1}^{n}(y_i - \beta_0)^2$

```{r}
#| echo: false
countdown::countdown(
  minutes = 2,
  play_sound = TRUE,
  color_border              = "#FEBC11",
  color_text                = "#003660",
  color_running_background  = "#003660",
  color_running_text        = "#FFFFFF",
  color_finished_background = "#FEBC11",
  color_finished_text       = "#003660",
  font_size = "2em",
)
```

:::
:::
:::
:::

---

---

## How do we pick $\beta_0$ and $\beta_1$? {.smaller}

We need a rule to pick the "best" $\widehat{\beta}_0$ and $\widehat{\beta}_1$. To do so, we want to make prediction errors small across all observations simultaneously.

The **residual** for observation $i$ is:
$$e_i = y_i - \widehat{y}_i$$

where $\widehat{y}_i$ is the **predicted value** $\widehat{\beta}_0 + \widehat{\beta}_1 x_i$

To find the best line, we minimize the **Residual Sum of Squares**:
$$\text{RSS} = \sum_{i=1}^n e_i^2 = \sum_{i=1}^n (y_i - \widehat{y}_i)^2$$

We square the residuals so positive and negative errors don't cancel, and so large errors are penalized more heavily.

---

## The Least Squares Criterion {.smaller}


From the previous slide, $e_i = y_i - \widehat{y}_i$. Since the candidate line predicts $\widehat{y}_i = \beta_0 + \beta_1 x_i$, substituting gives:

$$e_i = y_i - \beta_0 - \beta_1 x_i$$

So the **Residual Sum of Squares** is:

$$\text{RSS}(\beta_0, \beta_1) = \sum_{i=1}^{n} e_i^2 = \sum_{i=1}^{n}(y_i - \beta_0 - \beta_1 x_i)^2$$

**Goal:** choose $\widehat{\beta}_0$ and $\widehat{\beta}_1$ to minimize RSS.

---

## The Normal Equations {.smaller}

We have **2 unknowns** ($\beta_0$ and $\beta_1$) and need **2 equations** to pin them down.

Setting $\dfrac{\partial \text{RSS}}{\partial \beta_0} = 0$ and $\dfrac{\partial \text{RSS}}{\partial \beta_1} = 0$ each gives one equation — together they form a **system of two linear equations** called the **normal equations**.

Recall: $\text{RSS}(\beta_0, \beta_1) = \displaystyle\sum_{i=1}^{n}(y_i - \beta_0 - \beta_1 x_i)^2$

**Strategy:** differentiate RSS with respect to each parameter, set each derivative to zero, then use the chain rule and summation rule to simplify.

---

## The Normal Equations: Derivation {.smaller}

:::: {.columns style="font-size: 0.78em;"}

::: {.column width="50%"}

*Set $\partial\text{RSS}/\partial\beta_0 = 0$:*
$$\frac{\partial}{\partial \beta_0}\sum_{i=1}^{n}(y_i - \beta_0 - \beta_1 x_i)^2 = 0$$

::: {.fragment}
*Pass derivative through sum (summation rule):*
$$\sum_{i=1}^{n} \frac{\partial}{\partial \beta_0}(y_i - \beta_0 - \beta_1 x_i)^2 = 0$$
:::

::: {.fragment}
*Chain rule:*
$$\sum_{i=1}^{n} 2(y_i - \beta_0 - \beta_1 x_i) \cdot (-1) = 0$$
:::

::: {.fragment}
*Factor out $-2$, then divide both sides by $-2$:*
$$\sum_{i=1}^n (y_i - \beta_0 - \beta_1 x_i) = 0$$
:::

:::

::: {.column width="50%"}

*Set $\partial\text{RSS}/\partial\beta_1 = 0$:*
$$\frac{\partial}{\partial \beta_1}\sum_{i=1}^{n}(y_i - \beta_0 - \beta_1 x_i)^2 = 0$$

::: {.fragment}
*Pass derivative through sum (summation rule):*
$$\sum_{i=1}^{n} \frac{\partial}{\partial \beta_1}(y_i - \beta_0 - \beta_1 x_i)^2 = 0$$
:::

::: {.fragment}
*Chain rule:*
$$\sum_{i=1}^{n} 2(y_i - \beta_0 - \beta_1 x_i) \cdot (-x_i) = 0$$
:::

::: {.fragment}
*Factor out $-2$, then divide both sides by $-2$:*
$$\sum_{i=1}^n x_i(y_i - \beta_0 - \beta_1 x_i) = 0$$
:::

:::

::::


---


## Solving the Normal Equations: Step 1


**Solve equation 1 for $\widehat{\beta}_0$:**

$$\sum_{i=1}^n(y_i - \beta_0 - \beta_1 x_i) = 0 \;\Rightarrow\; n\bar{y} - n\beta_0 - n\beta_1\bar{x} = 0$$

$$\boxed{\widehat{\beta}_0 = \bar{y} - \widehat{\beta}_1\bar{x}}$$

---

## Solving the Normal Equations: Step 2 {.smaller}

**Substitute into equation 2 and solve for $\widehat{\beta}_1$:**

Start with equation 2 and distribute the sum:
$$\sum_{i=1}^n x_i y_i - \beta_0\sum_{i=1}^n x_i - \beta_1\sum_{i=1}^n x_i^2 = 0$$

Substitute $\beta_0 = \bar{y} - \beta_1\bar{x}$ and use $\sum x_i = n\bar{x}$:
$$\sum_{i=1}^n x_i y_i - (\bar{y} - \beta_1\bar{x})\,n\bar{x} - \beta_1\sum_{i=1}^n x_i^2 = 0$$

Expand the multiplication:
$$\sum_{i=1}^n x_i y_i - n\bar{x}\bar{y} + n\bar{x}^2\beta_1 - \beta_1\sum_{i=1}^n x_i^2 = 0$$

---

## Solving the Normal Equations: Step 2 (cont.) {.smaller}

Collect $\beta_1$ terms on the right:
$$\sum_{i=1}^n x_i y_i - n\bar{x}\bar{y} = \beta_1\!\left(\sum_{i=1}^n x_i^2 - n\bar{x}^2\right)$$

Divide both sides by $\left(\sum x_i^2 - n\bar{x}^2\right)$:

$$\boxed{\widehat{\beta}_1 = \frac{\displaystyle\sum_{i=1}^n x_i y_i - n\bar{x}\bar{y}}{\displaystyle\sum_{i=1}^n x_i^2 - n\bar{x}^2} = \frac{S_{xy}}{S_{xx}}}$$

---

## Equivalent Form of $\widehat{\beta}_1$ {.smaller}

The numerator and denominator have a cleaner equivalent form. Expanding $\sum(x_i - \bar{x})(y_i - \bar{y})$:

$$\sum(x_i - \bar{x})(y_i - \bar{y}) = \sum x_iy_i - \bar{y}\sum x_i - \bar{x}\sum y_i + n\bar{x}\bar{y}$$

Since $\sum x_i = n\bar{x}$ and $\sum y_i = n\bar{y}$:

$$= \sum x_iy_i - n\bar{x}\bar{y} - n\bar{x}\bar{y} + n\bar{x}\bar{y} = \sum x_iy_i - n\bar{x}\bar{y}$$

The same identity holds for the denominator: $\displaystyle\sum(x_i - \bar{x})^2 = \sum x_i^2 - n\bar{x}^2$


So the two forms are equivalent. We will use the simpler form on the next slide. 

---

## The OLS Estimators {.smaller}

Solving the normal equations yields:

$$\widehat{\beta}_1 = \frac{\displaystyle\sum_{i=1}^n (x_i - \bar{x})(y_i - \bar{y})}{\displaystyle\sum_{i=1}^n (x_i - \bar{x})^2} = \frac{S_{xy}}{S_{xx}}$$

$$\widehat{\beta}_0 = \bar{y} - \widehat{\beta}_1 \bar{x}$$

where:

$$S_{xy} = \sum_{i=1}^n (x_i - \bar{x})(y_i - \bar{y}) \quad \text{(co-variation of } x \text{ and } y \text{ around their means)}$$

$$S_{xx} = \sum_{i=1}^n (x_i - \bar{x})^2 \quad \text{(total spread of } x \text{ around its mean)}$$


---

## Connection to Correlation {.smaller}


The OLS slope has a useful rewrite in terms of correlation:

$$\widehat{\beta}_1 = r \cdot \frac{s_y}{s_x}$$

where $r$ is the sample correlation between $x$ and $y$, and $s_x$, $s_y$ are sample standard deviations.

::: {.incremental}
- The sign of $\widehat{\beta}_1$ always equals the sign of $r$
- When $r = 0$ $\rightarrow$ slope is 0 $\rightarrow$ no linear relationship
- When $|r| = 1$ $\rightarrow$ all points lie exactly on a line $\rightarrow$ perfect fit
:::

---

## Properties of the OLS Estimators {.smaller}

::: {style="font-size: 0.82em;"}
**Notation:** $S_{xx} = \displaystyle\sum_{i=1}^n(x_i - \bar{x})^2$ — total spread of the predictor; $\sigma^2$ — true (unknown) error variance; $\mathbb{E}(\cdot)$ — expected value; $\text{Var}(\cdot)$ — variance.
:::

::: {.incremental}
1. **Linear:** $\widehat{\beta}_0$ and $\widehat{\beta}_1$ are linear functions of $y_1, \ldots, y_n$

2. **Unbiased:** $\mathbb{E}(\widehat{\beta}_0) = \beta_0$ and $\mathbb{E}(\widehat{\beta}_1) = \beta_1$ — on average, the estimates recover the true parameters

3. **Known variances:**
$$\text{Var}(\widehat{\beta}_1) = \frac{\sigma^2}{S_{xx}}, \qquad \text{Var}(\widehat{\beta}_0) = \sigma^2\!\left(\frac{1}{n} + \frac{\bar{x}^2}{S_{xx}}\right)$$

4. **Gauss-Markov (BLUE):** OLS has the smallest variance among all linear unbiased estimators — no other such estimator can do better.
:::

---

## Property 1: Linearity of $\widehat{\beta}_1$ {.smaller}

::: {style="font-size: 0.82em;"}
$\widehat{\beta}_1$ is a **linear combination** $\sum a_i y_i$ where weights $a_i$ depend only on $x$. Starting from:
$\widehat{\beta}_1 = \dfrac{\sum(x_i - \bar{x})(y_i - \bar{y})}{S_{xx}}$

::: {.fragment}
**Step 1 — split the numerator:**
$$= \frac{\sum(x_i-\bar{x})y_i \;-\; \sum(x_i-\bar{x})\bar{y}}{S_{xx}}$$
:::

::: {.fragment}
**Step 2 — the second term is zero.** Since $\sum(x_i-\bar{x})=0$ (deviations from mean sum to zero) and $\bar{y}$ is a constant:
$$\sum(x_i-\bar{x})\bar{y} = \bar{y}\underbrace{\sum(x_i-\bar{x})}_{=\;0} = 0$$
:::

::: {.fragment}
**Step 3 — define weights** $a_i = (x_i - \bar{x})/S_{xx}$, which depend only on $x_i$:
$$\widehat{\beta}_1 = \frac{\sum(x_i-\bar{x})\,y_i}{S_{xx}} = \sum_{i=1}^n \underbrace{\frac{x_i - \bar{x}}{S_{xx}}}_{a_i} y_i = \sum_{i=1}^n a_i\,y_i \quad \checkmark$$
:::
:::

---

## Property 1: Linearity of $\widehat{\beta}_0$ {.smaller}

::: {style="font-size: 0.85em;"}
From the previous slide, $\widehat{\beta}_1 = \sum a_i y_i$. Now show $\widehat{\beta}_0 = \bar{y} - \widehat{\beta}_1\bar{x}$ is also linear.

**Step 1 — write $\bar{y}$ as a sum** (the sample mean is itself a linear combination):
$$\bar{y} = \sum_{i=1}^n \frac{1}{n}y_i$$

**Step 2 — substitute and collect** $y_i$ terms:
$$\widehat{\beta}_0 = \sum_{i=1}^n \frac{1}{n}y_i - \bar{x}\sum_{i=1}^n a_i y_i = \sum_{i=1}^n\!\underbrace{\left(\frac{1}{n} - a_i\bar{x}\right)}_{\text{weight for }y_i}y_i$$

The weights $\left(\tfrac{1}{n} - a_i\bar{x}\right)$ depend only on $x_i$, not $y_i$.

Both $\widehat{\beta}_0$ and $\widehat{\beta}_1$ are linear combinations of $y_1,\ldots,y_n$. 
:::

---

## Property 2: $\widehat{\beta}_1$ is Unbiased {.smaller}

::: {.callout-note title="Refresher: Expected Value"}
Key properties:

- $\mathbb{E}(c) = c$
- $\mathbb{E}(aY) = a\,\mathbb{E}(Y)$
- $\mathbb{E}(Y_1+Y_2)=\mathbb{E}(Y_1)+\mathbb{E}(Y_2)$
- $\mathbb{E}\!\left(\sum_i a_i Y_i\right) = \sum_i a_i\,\mathbb{E}(Y_i)$
:::

**Goal:** show $\mathbb{E}(\widehat{\beta}_1) = \beta_1$, i.e. the estimator is correct on average.

From the model, $y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$ with $\mathbb{E}(\varepsilon_i)=0$, so $\mathbb{E}(y_i) = \beta_0 + \beta_1 x_i$.

---

## Property 2: $\widehat{\beta}_1$ is Unbiased (cont.) {.smaller}

::: {style="font-size: 0.85em;"}

Since $a_i$ are constants, linearity of expectation lets us pull them out:
$$\mathbb{E}(\widehat{\beta}_1) = \mathbb{E}\!\left(\sum_{i=1}^n a_i y_i\right) = \sum_{i=1}^n a_i\,\mathbb{E}(y_i)$$

Substitute $\mathbb{E}(y_i) = \beta_0 + \beta_1 x_i$:
$$= \sum_{i=1}^n a_i(\beta_0 + \beta_1 x_i) = \beta_0\underbrace{\sum_{i=1}^n a_i}_{=\,0} + \beta_1\underbrace{\sum_{i=1}^n a_i x_i}_{=\,1}$$

**Why the sums equal 0 and 1:** recall $a_i = (x_i - \bar{x})/S_{xx}$

- $\sum a_i = \sum(x_i-\bar{x})/S_{xx} = 0$ since deviations from the mean sum to zero
- $\sum a_i x_i = \sum(x_i-\bar{x})x_i/S_{xx} = S_{xx}/S_{xx} = 1$

Therefore $\mathbb{E}(\widehat{\beta}_1) = \beta_0 \cdot 0 + \beta_1 \cdot 1 = \beta_1$. 

:::

---

## Property 2: $\widehat{\beta}_0$ is Unbiased {.smaller}

**Goal:** show $\mathbb{E}(\widehat{\beta}_0) = \beta_0$.

**Step 1 — find $\mathbb{E}(\bar{y})$:** expand using the model $\mathbb{E}(y_i) = \beta_0 + \beta_1 x_i$:

$$\mathbb{E}(\bar{y}) = \frac{1}{n}\sum_{i=1}^n \mathbb{E}(y_i) = \frac{1}{n}\sum_{i=1}^n(\beta_0 + \beta_1 x_i) = \frac{1}{n}(n\beta_0 + \beta_1 n\bar{x}) = \beta_0 + \beta_1\bar{x}$$

**Step 2 — apply linearity of expectation to $\widehat{\beta}_0 = \bar{y} - \widehat{\beta}_1\bar{x}$:** since $\bar{x}$ is a fixed constant,

$$\mathbb{E}(\widehat{\beta}_0) = \mathbb{E}(\bar{y}) - \bar{x}\,\mathbb{E}(\widehat{\beta}_1)$$

**Step 3 — substitute and cancel:** plug in $\mathbb{E}(\bar{y}) = \beta_0 + \beta_1\bar{x}$ and $\mathbb{E}(\widehat{\beta}_1) = \beta_1$:

$$\mathbb{E}(\widehat{\beta}_0) = (\beta_0 + \beta_1\bar{x}) - \bar{x}\,\beta_1 = \beta_0 $$

---

## Property 3: Variance of $\widehat{\beta}_1$ {.smaller}

**Goal:** derive the formula $\text{Var}(\widehat{\beta}_1) = \sigma^2 / S_{xx}$.

**Step 1 — apply the variance of a linear combination:** since $y_i$'s are independent with $\text{Var}(y_i)=\sigma^2$, and $a_i$ are fixed constants:

$$\text{Var}(\widehat{\beta}_1) = \text{Var}\!\left(\sum_{i=1}^n a_i y_i\right) \overset{\text{indep.}}{=} \sum_{i=1}^n a_i^2\,\text{Var}(y_i) = \sigma^2\sum_{i=1}^n a_i^2$$

Independence eliminates all cross-terms $a_ia_j\text{Cov}(y_i,y_j)$ for $i\neq j$.

---

## Property 3: Variance of $\widehat{\beta}_1$ (cont.) {.smaller}

**Step 2 — evaluate $\sum a_i^2$:** substitute $a_i = (x_i-\bar{x})/S_{xx}$ and use $\sum(x_i-\bar{x})^2 = S_{xx}$:

$$\sum_{i=1}^n a_i^2 = \sum_{i=1}^n \frac{(x_i-\bar{x})^2}{S_{xx}^2} = \frac{1}{S_{xx}^2}\underbrace{\sum_{i=1}^n(x_i-\bar{x})^2}_{=\,S_{xx}} = \frac{1}{S_{xx}}$$

**Step 3 — combine:**

$$\text{Var}(\widehat{\beta}_1) = \sigma^2 \cdot \frac{1}{S_{xx}} = \frac{\sigma^2}{S_{xx}} \qquad$$

---

## Property 3: Variance of $\widehat{\beta}_0$ {.smaller}

**Goal:** derive $\text{Var}(\widehat{\beta}_0) = \sigma^2\!\left(\tfrac{1}{n}+\tfrac{\bar{x}^2}{S_{xx}}\right)$.

**Step 1 — expand the variance.** Since $\widehat{\beta}_0 = \bar{y} - \widehat{\beta}_1\bar{x}$ and $\bar{x}$ is a fixed constant:

$$\text{Var}(\widehat{\beta}_0) = \text{Var}(\bar{y} - \widehat{\beta}_1\bar{x}) = \text{Var}(\bar{y}) + \bar{x}^2\,\text{Var}(\widehat{\beta}_1) - 2\bar{x}\,\text{Cov}(\bar{y},\widehat{\beta}_1)$$

**Step 2 — substitute known variances.** Using $\text{Var}(\bar{y}) = \sigma^2/n$ and $\text{Var}(\widehat{\beta}_1) = \sigma^2/S_{xx}$:

$$= \frac{\sigma^2}{n} + \frac{\bar{x}^2\sigma^2}{S_{xx}} - 2\bar{x}\,\text{Cov}(\bar{y},\widehat{\beta}_1)$$

**Step 3 — handle the covariance.** It can be shown that $\text{Cov}(\bar{y}, \widehat{\beta}_1) = 0$ (see [proof](https://en.wikipedia.org/wiki/Simple_linear_regression)), so the last term drops out:

$$\text{Var}(\widehat{\beta}_0) = \frac{\sigma^2}{n} + \frac{\bar{x}^2\sigma^2}{S_{xx}} = \sigma^2\!\left(\frac{1}{n}+\frac{\bar{x}^2}{S_{xx}}\right) \qquad$$

---

## Property 4: Gauss-Markov (BLUE) {.smaller}

::: {style="font-size: 0.85em;"}
**Goal:** OLS has the smallest variance among all linear unbiased estimators.

Let $b_1^* = \sum_{i=1}^n c_i y_i$ be any competing linear unbiased estimator of $\beta_1$.

**Step 1 — unbiasedness constraints.** We need $\mathbb{E}(b_1^*) = \beta_1$ for all $\beta_0, \beta_1$. Substitute $\mathbb{E}(y_i) = \beta_0 + \beta_1 x_i$:
$$\mathbb{E}(b_1^*) = \sum_{i=1}^n c_i(\beta_0 + \beta_1 x_i) = \beta_0\sum_{i=1}^n c_i + \beta_1\sum_{i=1}^n c_i x_i$$
For this to equal $\beta_1$ for *every* possible $\beta_0$ and $\beta_1$, we match coefficients:
$$\underbrace{\sum_{i=1}^n c_i}_{\text{coeff. of }\beta_0} = 0 \qquad \text{and} \qquad \underbrace{\sum_{i=1}^n c_i x_i}_{\text{coeff. of }\beta_1} = 1$$
:::

---

## Property 4: Gauss-Markov (BLUE) (cont.) {.smaller}

::: {style="font-size: 0.85em;"}
**Step 2 — decompose the weights.** Write $c_i = a_i + d_i$, where $a_i = (x_i-\bar{x})/S_{xx}$ are the OLS weights and $d_i = c_i - a_i$ is the extra piece.

Both $c_i$ and $a_i$ satisfy the same two constraints:

- From Step 1: $\sum c_i = 0$ and $\sum c_i x_i = 1$ (required for $b_1^*$ to be unbiased)
- From Property 1: $\sum a_i = 0$ and $\sum a_i x_i = 1$ (shown when we proved linearity of $\widehat{\beta}_1$)

Subtracting gives $\sum d_i = 0$ and $\sum d_i x_i = 0$.
:::

---

## Property 4: Gauss-Markov (BLUE) (cont.) {.smaller}

::: {style="font-size: 0.85em;"}
**Step 2 (cont.) — show $\sum a_i d_i = 0$.** We need this so the cross term vanishes when we expand $\sum(a_i + d_i)^2$ in Step 3 — without it we can't isolate $\text{Var}(\widehat{\beta}_1)$. Using $\sum d_i = 0$ and $\sum d_i x_i = 0$:
$$\sum_{i=1}^n a_i d_i = \frac{\sum(x_i-\bar{x})d_i}{S_{xx}} = \frac{\sum d_i x_i - \bar{x}\sum d_i}{S_{xx}} = \frac{0 - 0}{S_{xx}} = 0$$
:::

---

## Property 4: Gauss-Markov (BLUE) (cont.) {.smaller}

::: {style="font-size: 0.85em;"}
**Step 3 — compare variances.** Two key rules:

- $\text{Var}(cY) = c^2\text{Var}(Y)$ — a constant scales variance by its square
- $\text{Var}(Y_1 + Y_2) = \text{Var}(Y_1) + \text{Var}(Y_2)$ when $Y_1, Y_2$ are independent (cross terms $2c_ic_j\text{Cov}(y_i,y_j) = 0$)

Applying both to $b_1^* = \sum c_i y_i$ with independent $y_i$'s and $\text{Var}(y_i) = \sigma^2$:
$$\text{Var}(b_1^*) = \sum_{i=1}^n c_i^2\,\text{Var}(y_i) = \sigma^2\sum_{i=1}^n c_i^2$$
:::

---

## Property 4: Gauss-Markov (BLUE) (cont.) {.smaller}

::: {style="font-size: 0.85em;"}
Substitute $c_i = a_i + d_i$ and expand:

$$\sigma^2\sum(a_i + d_i)^2 = \sigma^2\!\left(\sum a_i^2 + 2\underbrace{\sum a_i d_i}_{=\;0} + \sum d_i^2\right) = \sigma^2\sum a_i^2 + \sigma^2\sum d_i^2$$

Recognizing $\sigma^2 \sum a_i^2 = \text{Var}(\widehat{\beta}_1)$:

$$\text{Var}(b_1^*) = \text{Var}(\widehat{\beta}_1) + \underbrace{\sigma^2 \sum d_i^2}_{\geq\;0} \;\geq\; \text{Var}(\widehat{\beta}_1) \qquad $$

The extra term $\sigma^2\sum d_i^2 \geq 0$ always — with equality only when $d_i = 0$ for all $i$, i.e. when $b_1^* = \widehat{\beta}_1$.
:::

---

## Practice: Computing OLS Estimates {.smaller}

A researcher records daily exercise hours ($x$) and wellness score ($y$, out of 15) for $n = 5$ students:

:::: {.columns}

::: {.column width="38%"}

| $i$ | $x_i$ | $y_i$ |
|:---:|:------:|:------:|
| 1 | 1 | 4 |
| 2 | 2 | 3 |
| 3 | 3 | 8 |
| 4 | 4 | 9 |
| 5 | 5 | 11 |

::: {style="font-size: 0.85em;"}
**Given:** $\bar{x} = 3$, $\;\bar{y} = 7$, $\;S_{xx} = 10$, $\;S_{xy} = 20$
:::

:::

::: {.column width="62%"}

**Compute:**

1. $\widehat{\beta}_1 = S_{xy} / S_{xx}$

2. $\widehat{\beta}_0 = \bar{y} - \widehat{\beta}_1\bar{x}$

3. Fitted values $\;\widehat{y}_i = \widehat{\beta}_0 + \widehat{\beta}_1 x_i$

4. Residuals $\;e_i = y_i - \widehat{y}_i$

```{r}
#| echo: false
countdown::countdown(
  minutes = 5,
  play_sound = TRUE,
  color_border              = "#FEBC11",
  color_text                = "#003660",
  color_running_background  = "#003660",
  color_running_text        = "#FFFFFF",
  color_finished_background = "#FEBC11",
  color_finished_text       = "#003660",
  font_size = "2em",
)
```

:::

::::

---

## Practice: Solution {.smaller}

:::: {.columns}

::: {.column width="48%"}

**Steps 1–2: Estimates**

$$\widehat{\beta}_1 = \frac{S_{xy}}{S_{xx}} = \frac{20}{10} = 2$$

$$\widehat{\beta}_0 = \bar{y} - \widehat{\beta}_1\bar{x} = 7 - 2(3) = 1$$

Fitted line: $\;\widehat{y} = 1 + 2x$

:::

::: {.column width="52%"}

**Steps 3–4: Fitted values and residuals**

| $i$ | $x_i$ | $y_i$ | $\widehat{y}_i$ | $e_i$ |
|:---:|:------:|:------:|:---------------:|:-----:|
| 1 | 1 | 4 | 3 | 1 |
| 2 | 2 | 3 | 5 | −2 |
| 3 | 3 | 8 | 7 | 1 |
| 4 | 4 | 9 | 9 | 0 |
| 5 | 5 | 11 | 11 | 0 |

:::

::::

---

## Summary {.smaller}

::: {.incremental}
- **Least squares:** OLS picks $\widehat{\beta}_0$ and $\widehat{\beta}_1$ to minimize $\text{RSS} = \sum(y_i - \widehat{y}_i)^2$
- **Normal equations** give closed-form solutions: $\widehat{\beta}_1 = S_{xy}/S_{xx}$, $\;\widehat{\beta}_0 = \bar{y} - \widehat{\beta}_1\bar{x}$
- **Residuals** $e_i = y_i - \widehat{y}_i$ measure how far each observation falls from the fitted line
- **Property 1 (Linearity):** both estimators are linear combinations of $y_1, \ldots, y_n$
- **Property 2 (Unbiasedness):** $\mathbb{E}(\widehat{\beta}_0) = \beta_0$ and $\mathbb{E}(\widehat{\beta}_1) = \beta_1$
- **Property 3 (Known variances):** $\text{Var}(\widehat{\beta}_1) = \sigma^2/S_{xx}$, $\;\text{Var}(\widehat{\beta}_0) = \sigma^2(1/n + \bar{x}^2/S_{xx})$
- **Property 4 (Gauss-Markov):** OLS is BLUE — smallest variance among all linear unbiased estimators
:::

---


## Questions? {background-color="#003660"}

::: {style="text-align: center; margin-top: 2em;"}
**Annie Adams**
[aradams@ucsb.edu](mailto:aradams@ucsb.edu)

Office Hours: Tuesday 10 – 11 AM

[Course Website](https://pstat-126.github.io/)
:::
