---
title: "Lecture 2: Fitting the Simple Linear Model"
subtitle: "PSTAT 126 — Linear Regression"
author: "Annie Adams"
institute: "UC Santa Barbara"
date: "Week 1 · Wednesday"
format:
  revealjs:
    theme: [default, ../styles.scss]
    slide-number: true
    preview-links: auto
    footer: "PSTAT 126 — Fall 2026 · UC Santa Barbara"
    logo: ""
    title-slide-attributes:
      data-background-color: "#003660"
---

## Review: The Simple Linear Model

From Monday, the simple linear regression model is:

$$y_i = \beta_0 + \beta_1 x_i + \varepsilon_i, \quad \varepsilon_i \overset{\text{iid}}{\sim} \mathcal{N}(0, \sigma^2)$$

- $\beta_1$ is a **comparison** between units — not a causal effect
- The fitted line $\hat{y} = \hat{\beta}_0 + \hat{\beta}_1 x$ summarizes the relationship

**Today:** How do we estimate $\hat{\beta}_0$, $\hat{\beta}_1$, and $\sigma$ from data?

---

## Learning Objectives

By the end of today you will be able to:

- Explain the least squares criterion and apply it conceptually
- State the OLS formulas for the slope and intercept
- Compute and interpret fitted values and residuals
- Interpret the residual standard deviation $\hat{\sigma}$ and $R^2$
- Explain what a standard error measures
- Use `lm()` in R to fit a simple linear regression and read its output

---

## How Do We Fit a Line? {background-color="#003660"}

---

## The Problem of Fitting

Given data $(x_1, y_1), (x_2, y_2), \ldots, (x_n, y_n)$, infinitely many lines pass through or near the points.

We need a **criterion** to select the "best" line.

**Intuition:** The best line should make prediction errors as small as possible across all observations simultaneously.

---

## Residuals: Measuring Prediction Error

For a candidate line with intercept $a$ and slope $b$, the **residual** for observation $i$ is:

$$e_i = y_i - \hat{y}_i = y_i - (a + b x_i)$$

- $\hat{y}_i = a + bx_i$: the **fitted value** (what the line predicts)
- $e_i > 0$: observed point is above the line
- $e_i < 0$: observed point is below the line

A good line has small residuals throughout.

---

## Ordinary Least Squares (OLS) {background-color="#003660"}

---

## The Least Squares Criterion

**Ordinary Least Squares (OLS)** chooses $\hat{\beta}_0$ and $\hat{\beta}_1$ to minimize the **Residual Sum of Squares (RSS)**:

$$\text{RSS}(a, b) = \sum_{i=1}^{n} e_i^2 = \sum_{i=1}^{n} (y_i - a - b x_i)^2$$

Why **square** the residuals?

::: {.incremental}
- Positive and negative errors don't cancel each other out
- Large errors are penalized more heavily than small ones
- The squared criterion has a clean, closed-form solution
:::

---

## Deriving the OLS Estimators

Minimize RSS by taking partial derivatives with respect to $a$ and $b$, setting them to zero, and solving:

$$\hat{\beta}_1 = \frac{\displaystyle\sum_{i=1}^n (x_i - \bar{x})(y_i - \bar{y})}{\displaystyle\sum_{i=1}^n (x_i - \bar{x})^2} = \frac{S_{xy}}{S_{xx}}$$

$$\hat{\beta}_0 = \bar{y} - \hat{\beta}_1 \bar{x}$$

::: {.callout-note}
The OLS line always passes through the point $(\bar{x},\, \bar{y})$.
:::

---

## A Useful Rewrite

The OLS slope can also be written as:

$$\hat{\beta}_1 = r \cdot \frac{s_y}{s_x}$$

where $r$ is the **sample correlation** between $x$ and $y$, and $s_x$, $s_y$ are sample standard deviations.

::: {.incremental}
- When $r = 0$: slope is 0 — no linear relationship
- When $|r| = 1$: all points lie exactly on a line
- The sign of $\hat{\beta}_1$ always equals the sign of $r$
:::

---

## Fitting the Election Model in R

```{r}
#| echo: true
#| eval: false
hibbs <- read.table("hibbs.dat", header = TRUE)
fit   <- lm(vote ~ growth, data = hibbs)
summary(fit)
```

```
Coefficients:
            Estimate Std. Error t value Pr(>|t|)
(Intercept)  46.248      1.623   28.50  < 2e-16 ***
growth        3.061      0.710    4.31  0.00076 ***

Residual standard error: 3.763 on 14 degrees of freedom
Multiple R-squared: 0.5700
```

---

## Reading the `lm()` Output {.smaller}

| Term | What it means |
|:---|:---|
| `(Intercept)` | $\hat{\beta}_0 = 46.2$: predicted vote share when growth = 0% |
| `growth` | $\hat{\beta}_1 = 3.1$: predicted change in vote share per 1% more income growth |
| `Std. Error` | Estimated standard error of each coefficient |
| `t value` | Coefficient estimate divided by its standard error |
| `Pr(>|t|)` | p-value for the hypothesis $H_0: \beta = 0$ |
| `Residual standard error` | $\hat{\sigma} = 3.76$: typical size of prediction errors |
| `Multiple R-squared` | $R^2 = 0.57$: fraction of variance in $y$ explained by the model |

---

## Residuals and Model Fit {background-color="#003660"}

---

## Accessing Residuals in R

```{r}
#| echo: true
#| eval: false
residuals(fit)   # e_i = y_i - y_hat_i
fitted(fit)      # y_hat_i = beta0_hat + beta1_hat * x_i

# Residual plot
plot(fitted(fit), residuals(fit),
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs. Fitted")
abline(h = 0, lty = 2, col = "gray50")
```

A well-fitted model has residuals that are centered at zero with no systematic pattern.

---

## Estimating $\sigma$

The true error variance $\sigma^2$ is unknown — we estimate it from the residuals:

$$\hat{\sigma}^2 = \frac{\text{RSS}}{n - 2} = \frac{\displaystyle\sum_{i=1}^n (y_i - \hat{y}_i)^2}{n - 2}$$

We divide by $n - 2$ — not $n$ — because we estimated **two parameters** ($\beta_0$ and $\beta_1$) from the data, using up 2 degrees of freedom.

---

## What Does $\hat{\sigma}$ Tell Us?

$\hat{\sigma}$ is the **residual standard deviation** — the typical size of prediction errors, on the scale of $y$.

For the election model: $\hat{\sigma} = 3.76$ percentage points.

**Interpretation:** Even with the best linear prediction using income growth, vote shares typically deviate from predictions by about **3.8 percentage points**.

::: {.callout-tip}
Compare $\hat{\sigma}$ to the spread of $y$ to assess predictive accuracy. A small $\hat{\sigma}$ relative to the range of $y$ indicates a useful model.
:::

---

## $R^2$: Proportion of Variance Explained

$$R^2 = 1 - \frac{\text{RSS}}{\text{TSS}} = 1 - \frac{\displaystyle\sum(y_i - \hat{y}_i)^2}{\displaystyle\sum(y_i - \bar{y})^2}$$

- **TSS** = Total Sum of Squares — variance in $y$ if we ignored $x$ entirely
- $R^2 = 0$: the model explains nothing; predicting $\bar{y}$ is just as good
- $R^2 = 1$: the model explains everything; residuals are all zero
- Election model: $R^2 = 0.57$ — income growth explains **57%** of the variation in vote share

::: {.callout-caution}
High $R^2$ does not mean the model is correct or causal — it only measures linear fit.
:::

---

## Standard Errors {background-color="#003660"}

---

## What is a Standard Error?

A **standard error** measures the precision of a sample estimate.

Imagine drawing many independent samples of size $n$ from the same population and computing $\hat{\beta}_1$ each time. The standard deviation of those estimates across samples is the **standard error** of $\hat{\beta}_1$.

$$\text{SE}(\hat{\beta}_1) = \frac{\hat{\sigma}}{\sqrt{S_{xx}}} = \frac{\hat{\sigma}}{\sqrt{\displaystyle\sum_{i=1}^n (x_i - \bar{x})^2}}$$

---

## What Makes Standard Errors Smaller?

$$\text{SE}(\hat{\beta}_1) = \frac{\hat{\sigma}}{\sqrt{\sum(x_i - \bar{x})^2}}$$

Standard errors are **smaller** (more precise estimates) when:

::: {.incremental}
- $\hat{\sigma}$ is small — less noise in the data
- Predictors $x$ are more spread out — more information about the slope
- Sample size $n$ is larger — more data overall
:::

---

## Confidence Intervals for Coefficients

A **95% confidence interval** for $\beta_1$ is:

$$\hat{\beta}_1 \pm t_{n-2,\,0.025} \cdot \text{SE}(\hat{\beta}_1)$$

For the election model ($n = 16$, $t_{14, 0.025} \approx 2.14$):

$$3.06 \pm 2.14 \times 0.71 \approx (1.54,\; 4.58)$$

**Interpretation:** We are 95% confident the true slope lies between 1.54 and 4.58 — a 1% increase in income growth is associated with between 1.5 and 4.6 more percentage points of vote share.

---

## Regression as Comparison {background-color="#003660"}

---

## Comparing Two Groups as Regression

Regression is a generalization of simpler statistical comparisons.

**Comparing a single mean:** regress $y$ on just a constant:
$$y_i = \mu + \varepsilon_i \quad \Longrightarrow \quad \hat{\mu} = \bar{y}$$

**Comparing two groups:** let $x_i \in \{0, 1\}$ be a binary group indicator:
$$y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$$

- $\hat{\beta}_0 = \bar{y}_{\text{group 0}}$
- $\hat{\beta}_1 = \bar{y}_{\text{group 1}} - \bar{y}_{\text{group 0}}$

This is algebraically equivalent to a two-sample $t$-test (assuming equal variances).

---

## Regression Generalizes the $t$-test

```{r}
#| echo: true
#| eval: false
# Suppose x is a binary 0/1 indicator (e.g., treatment vs. control)
t.test(y ~ x, var.equal = TRUE)

# Produces the same estimate as:
lm(y ~ x)
```

Regression is strictly more general — it can:

- Handle continuous predictors
- Adjust for additional variables simultaneously (multiple regression)
- Accommodate more than two groups via indicator coding

---

## Making Predictions

Given a new observation with predictor value $x^*$, the predicted outcome is:

$$\hat{y}^* = \hat{\beta}_0 + \hat{\beta}_1 x^*$$

**Example:** What vote share does the model predict when income growth = 2%?

$$\hat{y} = 46.2 + 3.1 \times 2 = 52.4\%$$

::: {.callout-warning}
## Extrapolation
Predicting far outside the range of the observed $x$ values is unreliable. The linear relationship may not hold beyond the data.
:::

---

## Brief Intro: Fake-Data Simulation

A powerful way to **validate** our fitting procedure: simulate data from a known model, then check whether we can recover the true parameters.

```{r}
#| echo: true
#| eval: false
set.seed(42)
n     <- 50
x     <- runif(n, min = 0, max = 10)

# True (known) parameters
beta0 <- 2
beta1 <- 3
sigma <- 5

y   <- beta0 + beta1 * x + rnorm(n, mean = 0, sd = sigma)
fit <- lm(y ~ x)
coef(fit)  # should be close to (2, 3)
```

If estimates are far from the truth on simulated data, something is wrong with our procedure.

---


## Coefficients as Comparisons {background-color="#003660"}

---

## Regression Coefficients Describe Comparisons

A regression slope $\hat{\beta}_1$ compares groups of units that differ by 1 unit in $x$.

$\hat{\beta}_1$ is the estimated average difference in $y$ between two otherwise-similar observations where $x$ differs by 1 unit.

::: {.callout-important}
## Correlation $\neq$ Causation
Regression coefficients describe **associations** in data. They do not, by themselves, establish causal relationships.
:::

---

## Why "Effect" Language Can Mislead

Consider: *"Each year of education **increases** income by $5,000."*

This phrasing implies a causal mechanism — that giving someone more education *would* change their income by $5,000.

What the regression actually estimates:

> "People with one more year of education earn, **on average**, $5,000 more than people with one fewer year of education."

The difference matters when drawing policy conclusions.

---

## A Better Framing

::: {.columns}
::: {.column width="48%"}
**Avoid:**

*"Education increases income by $5,000."*
:::
::: {.column width="48%"}
**Prefer:**

*"Comparing people who differ by one year of education, those with more education earn $5,000 more on average."*
:::
:::

Regression describes **comparisons across units in the data**. Causal claims require additional study design considerations beyond this course.

---

## Summary {.smaller}

::: {.incremental}
- **OLS** chooses $\hat{\beta}_0$ and $\hat{\beta}_1$ by minimizing $\text{RSS} = \sum(y_i - \hat{y}_i)^2$
- Closed-form solution: $\hat{\beta}_1 = S_{xy}/S_{xx}$, $\;\hat{\beta}_0 = \bar{y} - \hat{\beta}_1 \bar{x}$
- **Residuals** $e_i = y_i - \hat{y}_i$ measure prediction errors
- **$\hat{\sigma}$** is the residual SD — typical prediction error — estimated by $\sqrt{\text{RSS}/(n-2)}$
- **$R^2$** is the proportion of variance in $y$ explained by $x$ — useful but limited
- **Standard errors** quantify uncertainty in coefficient estimates
- Regression with a binary predictor generalizes the two-sample $t$-test
- **Fake-data simulation** is a principled tool for validating our fitting procedure
:::

---

## Looking Ahead

**Week 2:** Multiple Linear Regression

- Adding more predictors: $y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots + \varepsilon$
- Interpreting partial slopes — "holding other predictors constant"
- Indicator variables and categorical predictors

**Reading:** *Regression and Other Stories*, Chapters 7.2–7.3 & Chapter 8

---

## Questions? {background-color="#003660"}

::: {style="text-align: center; margin-top: 2em;"}
**Annie Adams**
[aradams@ucsb.edu](mailto:aradams@ucsb.edu)

Office Hours: Tuesday 10 – 11 AM

[Course Website](https://pstat-126.github.io/)
:::
