---
title: "Lecture 4: Residuals and Estimating Error Variance"
subtitle: "PSTAT 126: Linear Regression"
author: "Annie Adams"
institute: "UC Santa Barbara"
date: "10/7/2026"
format:
  revealjs:
    theme: [default, ../styles.scss]
    slide-number: true
    preview-links: auto
    footer: "PSTAT 126 — Fall 2026 · UC Santa Barbara"
    logo: ""
    chalkboard: true
    title-slide-attributes:
      data-background-color: "#003660"
---

## Course Reminders 
- Labs are due Mondays by 11:59 PM
- HW 1 released today and due next Saturday, October 17th at 11:59 PM
- Quiz 1 on Monday in class 


## Review: OLS Estimators

From last time, the OLS estimators are:

$$\widehat{\beta}_1 = \frac{S_{xy}}{S_{xx}}, \qquad \widehat{\beta}_0 = \bar{y} - \widehat{\beta}_1\bar{x}$$

They are **unbiased**, **linear**, and the **B**est **L**inear **U**nbiased **E**stimator (Gauss-Markov).

**Today:** What do the residuals tell us, and how do we estimate the unknown $\sigma^2$?

---

## Learning Objectives

By the end of today you will be able to:

- Define residuals and explain what they measure
- State and prove the four properties of OLS residuals
- Define RSS and explain its role in regression
- Compute and interpret $\widehat{\sigma}^2$ and $\widehat{\sigma}$

---

## Residuals {background-color="#003660"}

---

## What are Residuals? {.smaller}

After fitting the model, the **residual** for observation $i$ is:

$$e_i = y_i - \widehat{y}_i = y_i - \widehat{\beta}_0 - \widehat{\beta}_1 x_i$$

Residuals measure how far each observation falls from the fitted line:

- $e_i > 0$: observed value is **above** the fitted line
- $e_i < 0$: observed value is **below** the fitted line
- $e_i = 0$: observation falls **exactly** on the fitted line

::: {.callout-note}
Residuals $e_i$ are the observable counterparts to the unobservable errors $\varepsilon_i$. We use them to check model assumptions and estimate $\sigma^2$.
:::

---

## Properties of OLS Residuals {.smaller}


1. $\displaystyle\sum_{i=1}^n e_i = 0$ — residuals sum to zero
2. $\displaystyle\sum_{i=1}^n \widehat{y}_i = \sum_{i=1}^n y_i$ — sum of fitted values equals sum of observed values
3. $\displaystyle\sum_{i=1}^n x_i e_i = 0$ — residuals are uncorrelated with the predictor
4. $\displaystyle\sum_{i=1}^n \widehat{y}_i e_i = 0$ — residuals are uncorrelated with the fitted values
5. The fitted line always passes through $(\bar{x}, \bar{y})$


---

## Estimating $\sigma^2$ {background-color="#003660"}

---

## Residual Sum of Squares {.smaller}

The **Residual Sum of Squares (RSS)** measures the total unexplained variation:

$$\text{RSS} = \sum_{i=1}^n e_i^2 = \sum_{i=1}^n (y_i - \widehat{y}_i)^2$$

- A smaller RSS means the line fits the data more closely
- RSS = 0 only if every point lies exactly on the fitted line


---

## Estimating $\sigma^2$ {.smaller}

The error variance $\sigma^2 = \text{Var}(\varepsilon_i)$ is unknown. We estimate it using the residuals.

A natural starting point is the average squared residual, but we must divide by $n - 2$, not $n$:

$$\widehat{\sigma}^2 = \text{MSE} = \frac{\text{RSS}}{n - 2} = \frac{\sum_{i=1}^n e_i^2}{n-2}$$

This quantity is also called the **Mean Square Error (MSE)**.

**Why $n - 2$?** Estimating $\widehat{\beta}_0$ and $\widehat{\beta}_1$ uses 2 degrees of freedom since two parameters are being used.

It can be shown that $\mathbb{E}(\widehat{\sigma}^2) = \sigma^2$, so $\widehat{\sigma}^2$ is an **unbiased** estimator of $\sigma^2$.

---



## Proof: Steps 1 & 2 {.smaller}

**Goal:** Show $\mathbb{E}(\text{MSE}) = \sigma^2$, i.e., $\mathbb{E}(\text{RSS}) = (n-2)\sigma^2$.

::: {style="font-size: 0.85em;"}
**Step 1 — rewrite RSS.** Substitute $\widehat{\beta}_0 = \bar{y} - \widehat{\beta}_1\bar{x}$ to center the residuals:

<div style="height: 6em;"></div>

**Step 2 — expand the square.** Use $\sum(y_i-\bar{y})(x_i-\bar{x}) = \widehat{\beta}_1 S_{xx}$, then take expectations:

<div style="height: 5em;"></div>


:::

---

## Proof: Step 3 — Piece 1 {.smaller}

::: {style="font-size: 0.85em;"}
**Evaluate Piece 1:** $\mathbb{E}\!\left(\sum(y_i - \bar{y})^2\right)$

Substitute $y_i - \bar{y} = \beta_1(x_i - \bar{x}) + (\varepsilon_i - \bar{\varepsilon})$ and expand the square:

<div style="height: 12em;"></div>

::: {.callout-note title="Recall: Model Assumption"}
$\mathbb{E}(\varepsilon_i) = 0$ is a model assumption (errors are mean-zero). Then by linearity of expectation: $\mathbb{E}(\bar{\varepsilon}) = \frac{1}{n}\sum_{i=1}^n \mathbb{E}(\varepsilon_i) = 0$.
:::

:::

---

## Proof: Step 3 — Piece 2 {.smaller}

::: {style="font-size: 0.85em;"}
**Evaluate Piece 2:** $\mathbb{E}\!\left(\widehat{\beta}_1^2\, S_{xx}\right)$

Since $S_{xx}$ is non-random, this reduces to finding $\mathbb{E}(\widehat{\beta}_1^2)$.

Use the identity: $\mathbb{E}(Z^2) = \text{Var}(Z) + \left[\mathbb{E}(Z)\right]^2$

<div style="height: 7em;"></div>


:::

---

## Proof: Step 4 — Combine {.smaller}

::: {style="font-size: 0.85em;"}
We now have both pieces. Plug them into $\mathbb{E}(\text{RSS}) = \text{Piece 1} - \text{Piece 2}$:






<div style="height: 6em;"></div>



<div style="height: 3em;"></div>

$\mathbb{E}(\text{MSE}) = \mathbb{E}\!\left(\dfrac{\text{RSS}}{n-2}\right) =$
:::

---

## Proof: Steps 1 & 2 (Solution) {.smaller}

::: {style="font-size: 0.85em;"}
**Step 1 — rewrite RSS.** Substitute $\widehat{\beta}_0 = \bar{y} - \widehat{\beta}_1\bar{x}$:

$$\text{RSS} = \sum(y_i - \widehat{y}_i)^2 = \sum\!\left[(y_i - \bar{y}) - \widehat{\beta}_1(x_i - \bar{x})\right]^2$$

**Step 2 — expand the square.** Since $\sum(y_i-\bar{y})(x_i-\bar{x}) = \widehat{\beta}_1 S_{xx}$:

$$= \sum(y_i-\bar{y})^2 - 2\widehat{\beta}_1^2 S_{xx} + \widehat{\beta}_1^2 S_{xx} = \sum(y_i-\bar{y})^2 - \widehat{\beta}_1^2 S_{xx}$$

:::

---

## Proof: Step 3 — Piece 1 (Solution) {.smaller}

::: {style="font-size: 0.85em;"}
**Evaluate Piece 1:** $\mathbb{E}\!\left(\sum(y_i - \bar{y})^2\right)$

Substitute $y_i - \bar{y} = \beta_1(x_i-\bar{x}) + (\varepsilon_i - \bar{\varepsilon})$ and expand:

$$\sum(y_i-\bar{y})^2 = \beta_1^2 S_{xx} + 2\beta_1\!\sum(x_i-\bar{x})(\varepsilon_i-\bar{\varepsilon}) + \sum(\varepsilon_i-\bar{\varepsilon})^2$$

Taking expectations — the cross term vanishes since $\mathbb{E}(\varepsilon_i - \bar\varepsilon) = 0$:

$$\mathbb{E}\!\left(\sum(y_i-\bar{y})^2\right) = \beta_1^2 S_{xx} + (n-1)\sigma^2$$


:::

---

## Proof: Step 3 — Piece 2 (Solution) {.smaller}

::: {style="font-size: 0.85em;"}
**Evaluate Piece 2:** $\mathbb{E}\!\left(\widehat{\beta}_1^2\, S_{xx}\right)$

Since $S_{xx}$ is non-random: $\mathbb{E}(\widehat{\beta}_1^2 S_{xx}) = S_{xx} \cdot \mathbb{E}(\widehat{\beta}_1^2)$.

Apply $\mathbb{E}(Z^2) = \text{Var}(Z) + [\mathbb{E}(Z)]^2$ with $Z = \widehat{\beta}_1$:

$$\mathbb{E}(\widehat{\beta}_1^2) = \text{Var}(\widehat{\beta}_1) + \left[\mathbb{E}(\widehat{\beta}_1)\right]^2 = \frac{\sigma^2}{S_{xx}} + \beta_1^2$$

Therefore: $\mathbb{E}(\widehat{\beta}_1^2 S_{xx}) = \sigma^2 + \beta_1^2 S_{xx}$



:::

---

## Proof: Step 4 — Combine (Solution) {.smaller}

::: {style="font-size: 0.85em;"}
Now plug both pieces in:

$$\mathbb{E}(\text{RSS}) = \Big[\beta_1^2 S_{xx} + (n-1)\sigma^2\Big] - \Big[\sigma^2 + \beta_1^2 S_{xx}\Big]$$

The $\beta_1^2 S_{xx}$ terms cancel, leaving:

$$= (n-1)\sigma^2 - \sigma^2 = (n-2)\sigma^2$$

$$\mathbb{E}(\text{MSE}) = \mathbb{E}\!\left(\frac{\text{RSS}}{n-2}\right) = \frac{(n-2)\sigma^2}{n-2} = \sigma^2 \qquad $$
:::

---

## The Residual Standard Deviation {.smaller}

Taking the square root gives the **residual standard deviation**:

$$\widehat{\sigma} = \sqrt{\widehat{\sigma}^2} = \sqrt{\frac{\text{RSS}}{n-2}}$$

**Interpretation:** $\widehat{\sigma}$ is the typical size of a prediction error, on the scale of $y$.

::: {.callout-tip}
Compare $\widehat{\sigma}$ to the spread of $y$ to gauge model usefulness. A small $\widehat{\sigma}$ relative to the range of $y$ indicates a good fit.
:::

---

## Standard Errors of the Estimators {.smaller}

Since $\sigma^2$ is unknown, we replace it with $\widehat{\sigma}^2$ in the variance formulas:

$$\widehat{\text{Var}}(\widehat{\beta}_1) = \frac{\widehat{\sigma}^2}{S_{xx}}, \qquad \widehat{\text{Var}}(\widehat{\beta}_0) = \widehat{\sigma}^2\!\left(\frac{1}{n} + \frac{\bar{x}^2}{S_{xx}}\right)$$

The **standard errors** are the square roots:

$$\text{SE}(\widehat{\beta}_1) = \frac{\widehat{\sigma}}{\sqrt{S_{xx}}}, \qquad \text{SE}(\widehat{\beta}_0) = \widehat{\sigma}\sqrt{\frac{1}{n} + \frac{\bar{x}^2}{S_{xx}}}$$

Standard errors quantify the uncertainty in each estimate — they are the basis for confidence intervals and hypothesis tests (coming soon).

---

## Practice Problem {.smaller}

Five weather stations record elevation $X$ (km) and average annual temperature $Y$ (°C), with $n=5$, $\bar{x}=3$, $\bar{y}=12$, $S_{xx}=10$, $S_{xy}=-30$:

:::: {.columns}

::: {.column width="52%"}

| $i$ | $x_i$ | $y_i$ | $\widehat{y}_i$ | $e_i$ |
|:---:|:------:|:------:|:---------------:|:-----:|
| 1 | 1 | 19 | | |
| 2 | 2 | 13 | | |
| 3 | 3 | 13 | | |
| 4 | 4 | 9 | | |
| 5 | 5 | 6 | | |

:::

::: {.column width="48%"}

**Compute & Interpret:**

1. Write the formula for $\widehat{\text{Temperature}}$
2. Interpret $\widehat{\beta}_1$ in context
3. Is $\widehat{\beta}_0$ meaningful here?
4. Fill in $\widehat{y}_i$ and $e_i$; verify $\sum e_i = 0$
5. Predict temperature at $x = 2.5$ km
6. Compute RSS, $\widehat{\sigma}^2$, $\widehat{\sigma}$, and $\text{SE}(\widehat{\beta}_1)$; interpret $\widehat{\sigma}$

```{r}
#| echo: false
countdown::countdown(
  minutes = 10,
  play_sound = TRUE,
  color_border              = "#FEBC11",
  color_text                = "#003660",
  color_running_background  = "#003660",
  color_running_text        = "#FFFFFF",
  color_finished_background = "#FEBC11",
  color_finished_text       = "#003660",
  font_size = "2em",
)
```

:::

::::

---


## Fitting Linear Regression in R {background-color="#003660"}

---

## The `lm()` Function {.smaller}

In R, we fit a linear regression model using `lm()`:

```r
lm(y ~ x, data = your_data)
```

- `y` is the **response** variable
- `x` is the **predictor** variable
- The `~` reads as "is modeled as a function of"
- The result is an object we can pass to `summary()`, `fitted()`, `residuals()`, etc.

---

## Example: Air Quality in New York {.smaller}

The built-in `airquality` dataset records daily air quality measurements in New York (May–September 1973).

- **X**: Temperature (°F)
- **Y**: Ozone concentration (ppb)

```{r}
#| echo: true
aq <- na.omit(airquality[, c("Temp", "Ozone")])

plot(aq$Temp, aq$Ozone,
     xlab = "Temperature (°F)",
     ylab = "Ozone (ppb)")
```

---

## Fitting the Model {.smaller}

```{r}
#| echo: true
fit <- lm(Ozone ~ Temp, data = aq)
summary(fit)
```

---

## Understanding `summary()` Output {.smaller}

```{r}
#| echo: false
summary(fit)
```

::: {.fragment .absolute top="322px" left="120px" width="88px" height="70px" fragment-index=1 style="border:3px solid #E8A838; border-radius:4px; background:rgba(232,168,56,0.2);"}
:::
::: {.fragment .absolute top="312px" left="595px" fragment-index=1 style="font-size:0.58em; color:#7D5A00; font-weight:bold; border-left:3px solid #E8A838; padding-left:8px; line-height:1.5;"}
**Estimates**
$\widehat\beta_0=-147.0,\;\widehat\beta_1=2.43$
:::

::: {.fragment .absolute top="322px" left="210px" width="95px" height="70px" fragment-index=2 style="border:3px solid #5BA4CF; border-radius:4px; background:rgba(91,164,207,0.2);"}
:::
::: {.fragment .absolute top="356px" left="595px" fragment-index=2 style="font-size:0.58em; color:#1A5276; font-weight:bold; border-left:3px solid #5BA4CF; padding-left:8px; line-height:1.5;"}
**Std. Errors**
SE($\widehat\beta_0$), SE($\widehat\beta_1$)
:::

::: {.fragment .absolute top="445px" left="0px" width="570px" height="26px" fragment-index=3 style="border:3px solid #58A867; border-radius:4px; background:rgba(88,168,103,0.2);"}
:::
::: {.fragment .absolute top="428px" left="595px" fragment-index=3 style="font-size:0.58em; color:#256B38; font-weight:bold; border-left:3px solid #58A867; padding-left:8px; line-height:1.5;"}
$\widehat\sigma=23.71$ — residual std. deviation
:::

::: {.fragment .absolute top="473px" left="0px" width="570px" height="23px" fragment-index=4 style="border:3px solid #C0392B; border-radius:4px; background:rgba(192,57,43,0.2);"}
:::
::: {.fragment .absolute top="461px" left="595px" fragment-index=4 style="font-size:0.58em; color:#922B21; font-weight:bold; border-left:3px solid #C0392B; padding-left:8px;"}
$R^2=0.488$ — variation explained by model
:::

---

## Interpreting the Output 
::: {style="font-size: 0.55em;"}

From the output above, we read off the key values directly:

::: {.callout-note title="Air Quality Example"}
We have $\widehat{\beta}_0 = -147.65$, $\widehat{\beta}_1 = 2.44$, and $\widehat{\sigma} = 23.71$.
Thus $\widehat{\sigma}^2 = 23.71^2 \approx 562.2$.

Fitted model: $\widehat{\text{Ozone}} = -147.65 + 2.44 \times \text{Temp}$
:::

**The intercept is included by default.** Use formula shortcuts to change this:

| Formula | Model |
|:---|:---|
| `lm(y ~ x)` | $Y_i = \beta_0 + \beta_1 X_i + \varepsilon_i$ — intercept included |
| `lm(y ~ x - 1)` | $Y_i = \beta_1 X_i + \varepsilon_i$ — no intercept (line through origin) |
| `lm(y ~ 1)` | $Y_i = \beta_0 + \varepsilon_i$ — intercept only (estimates the mean of $Y$) |
:::
---

## Adding the Fitted Line {.smaller}

Use `abline()` to overlay the OLS regression line:

```{r}
#| echo: true
plot(aq$Temp, aq$Ozone,
     xlab = "Temperature (°F)",
     ylab = "Ozone (ppb)")
abline(fit, col = "red")
```

---

## Extracting Key Quantities {.smaller}

Once a model is fit, R gives us easy access to the components we have been computing by hand:

```{r}
#| echo: true
coef(fit)        # beta_0 hat and beta_1 hat
fitted(fit)      # y-hat values
residuals(fit)   # e_i = y_i - y-hat_i
sigma(fit)       # sigma-hat (residual standard deviation)
```

::: {.callout-tip}
`sigma()` returns $\widehat{\sigma} = \sqrt{\text{RSS}/(n-2)}$ — the same quantity we computed by hand today.
:::

---

## Summary {.smaller}

::: {.incremental}
- **Residuals** $e_i = y_i - \widehat{y}_i$ are observed proxies for the unobservable errors $\varepsilon_i$
- **Five properties** of OLS residuals follow from the normal equations: $\sum e_i = 0$, $\sum \widehat{y}_i = \sum y_i$, $\sum x_i e_i = 0$, $\sum \widehat{y}_i e_i = 0$, and the line passes through $(\bar{x}, \bar{y})$
- **RSS** $= \sum e_i^2$ measures total unexplained variation — OLS minimizes it by construction
- **$\widehat{\sigma}^2 = \text{RSS}/(n-2)$** is an unbiased estimate of $\sigma^2$; dividing by $n-2$ corrects for two constraints placed on the residuals
- **$\widehat{\sigma}$** is the residual standard deviation — the typical prediction error on the scale of $y$
- **Standard errors** replace $\sigma$ with $\widehat{\sigma}$ in the variance formulas — the foundation for inference
- Use `lm()` in R to fit the model and `summary()`, `residuals()`, `sigma()` to extract results
:::

---

## Questions? {background-color="#003660"}

::: {style="text-align: center; margin-top: 2em;"}
**Annie Adams**
[aradams@ucsb.edu](mailto:aradams@ucsb.edu)

Office Hours: Tuesday 10 – 11 AM on [Zoom](https://ucsb.zoom.us/j/81194403106?pwd=sQo3HO4iaAnzcY0apRmIyAdC1KbVfT.1)


[Course Website](https://pstat-126.github.io/)
:::
