---
title: "Lecture 1: What is Regression?"
subtitle: "PSTAT 126 — Linear Regression"
author: "Annie Adams"
institute: "UC Santa Barbara"
date: "09/28/2026"
format:
  revealjs:
    theme: [default, ../styles.scss]
    slide-number: true
    preview-links: auto
    footer: "PSTAT 126 — Fall 2026 · UC Santa Barbara"
    logo: ""
    title-slide-attributes:
      data-background-color: "#003660"
---

```{r}
#| include: false
if (!requireNamespace("modeldata", quietly = TRUE)) install.packages("modeldata")
library(modeldata)
library(ggplot2)
data(ames)
```

## Learning Objectives

By the end of today you will be able to:

- Explain the historical origin of "regression" and what regression to the mean means
- Describe what a regression model is and what questions it can answer
- Write the simple linear regression model and identify its components
- Interpret a regression coefficient as a **comparison**, not a causal effect

---

## Historical Origins of Regression {background-color="#003660"}

---

## Francis Galton and Heredity

In the **1880s**, Francis Galton studied the relationship between parents' heights and their children's heights.

**His finding:** Tall parents tend to have taller children who are *shorter than the parents*. Short parents tend to have shorter children who are *taller than the parents*.

Children's heights "regressed" back toward the population mean.

This is the origin of the word **"regression."**


---

## What is a Statistical Model? {background-color="#003660"}

---

## What is a Statistical Model?
A **statistical model** is a way of describing real-world phenomena using mathematical relationships. It takes data as input and creates a structured way of understanding how variables relate to one another, and how those relationships can be used to predict an outcome. 





$$y = f(x) + \varepsilon$$

- $y$: outcome of interest (response variable)
- $x$: predictor (explanatory variable)
- $f(x)$: systematic component — what we can explain
- $\varepsilon$: random error — irreducible noise

**Regression** is the most widely used class of statistical models.

---

## Why Regression?

Regression lets us answer questions like:

- How does income relate to years of education?
- Does economic growth predict election outcomes?
- What is the average difference in test scores between two groups, *holding other factors constant*?
- Given a new observation, what do we predict for $y$?

---

## Forms of Regression Models {.smaller}

The regression framework is flexible enough to handle many data structures:


| Model Type | Form |
|:---|:---|
| Simple linear | $y = \beta_0 + \beta_1 x + \varepsilon$ |
| Multiple linear | $y = \beta_0 + \beta_1 x_1 + \cdots + \beta_p x_p + \varepsilon$ |
| Log-transformed | $\log(y) = \beta_0 + \beta_1 x + \varepsilon$ |
| With interactions | $y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_1 x_2 + \varepsilon$ |
| Logistic | $\log\!\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 x$ |

---

## A Motivating Example {background-color="#003660"}

---

## Predicting House Prices

**Question:** Can the size of a house predict its sale price?

Data: 2,930 home sales in Ames, Iowa (De Cock 2011)

- **Outcome** $y$: sale price ($1,000s)
- **Predictor** $x$: above-ground living area (square feet)


:::{.fragment}
::: callout-tip
## Turn & Talk

:::{.nonincremental}
::: {style="font-size: 30px"}
What do you think? How strong should this relationship be? Do you think it will be positive or negative? 

```{r}
#| echo: false
countdown::countdown(
  minutes = 1,
  play_sound = TRUE,
  color_border              = "#FEBC11",
  color_text                = "#003660",
  color_running_background  = "#003660",
  color_running_text        = "#FFFFFF",
  color_finished_background = "#FEBC11",
  color_finished_text       = "#003660",
  font_size = "2em",
)
```

:::
:::
:::
:::

---

## The Ames Housing Data

```{r}
#| echo: false
#| eval: true
#| fig-width: 8
#| fig-height: 4.5
#| fig-align: center
ggplot(ames, aes(x = Gr_Liv_Area, y = Sale_Price / 1000)) +
  geom_point(alpha = 0.3, size = 1.5) +
  labs(
    x = "Above-ground living area (sq ft)",
    y = "Sale price ($1,000s)"
  ) +
  theme_minimal()
```

What patterns do you notice in the scatter plot?

---

## The Simple Linear Regression Model {background-color="#003660"}

---

## The Model

The **simple linear regression model** relates one predictor to a continuous outcome:

$$y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$$

- $y_i$: observed outcome for unit $i$
- $x_i$: predictor for unit $i$
- $\beta_0$(**intercept**):  expected value of $y$ when $x = 0$
- $\beta_1$(**slope**): expected change in $y$ per one-unit increase in $x$
- $\varepsilon_i \overset{\text{iid}}{\sim} \mathcal{N}(0, \sigma^2)$: random error

---

## Interpreting Slope and Intercept

For the Ames housing model, fitting gives:
$$\widehat{\text{price}} = 13.3 + 0.111 \times \text{sqft}$$

::: {.incremental}
- **Intercept** ($\hat{\beta}_0 = 13.3$): a house with 0 sq ft is predicted to sell for $13,300 — not meaningful on its own, but a necessary mathematical anchor

- **Slope** ($\hat{\beta}_1 = 0.111$): each additional square foot of living area is associated with $111 more in sale price, *on average*
:::

---

## Adding the Fitted Line

```{r}
#| echo: false
#| eval: true
ggplot(ames, aes(x = Gr_Liv_Area, y = Sale_Price / 1000)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#003660") +
  labs(
    x = "Above-ground living area (sq ft)",
    y = "Sale price ($1,000s)",
    title = "Sale Price ~ Living Area (Ames, Iowa)"
  ) +
  theme_minimal()
```

The fitted line $\hat{y} = \hat{\beta}_0 + \hat{\beta}_1 x$ is our best linear summary of the relationship.

---

## Coefficients as Comparisons {background-color="#003660"}

---

## Regression Coefficients Describe Comparisons

A regression slope $\hat{\beta}_1$ compares groups of units that differ by 1 unit in $x$.

$\hat{\beta}_1$ is the estimated average difference in $y$ between two otherwise-similar observations where $x$ differs by 1 unit.

::: {.callout-important}
## Correlation $\neq$ Causation
Regression coefficients describe **associations** in data. They do not, by themselves, establish causal relationships.
:::

---

## Why "Effect" Language Can Mislead

Consider: *"Each year of education **increases** income by $5,000."*

This phrasing implies a causal mechanism — that giving someone more education *would* change their income by $5,000.

What the regression actually estimates:

> "People with one more year of education earn, **on average**, $5,000 more than people with one fewer year of education."

The difference matters when drawing policy conclusions.

---

## A Better Framing

::: {.columns}
::: {.column width="48%"}
**Avoid:**

*"Education increases income by $5,000."*
:::
::: {.column width="48%"}
**Prefer:**

*"Comparing people who differ by one year of education, those with more education earn $5,000 more on average."*
:::
:::

Regression describes **comparisons across units in the data**. Causal claims require additional study design considerations beyond this course.

---

## Summary {.smaller}

::: {.incremental}
- Regression originated from Galton's 19th-century study of heredity
- **Regression to the mean** is a statistical phenomenon driven by random error, not a causal mechanism
- Regression models the relationship between $y$ and $x$: $y = \beta_0 + \beta_1 x + \varepsilon$
- The **slope** $\beta_1$ describes how $y$ changes with $x$, on average — a comparison between units
- The **intercept** $\beta_0$ is the predicted value of $y$ when $x = 0$
- Coefficients describe **comparisons**, not causal effects — language matters
:::

---

## Looking Ahead

**Wednesday:** How do we actually *fit* a regression model to data?

- Ordinary Least Squares (OLS) — minimizing prediction errors
- Residuals and residual standard deviation $\hat{\sigma}$
- Standard errors and uncertainty in estimates
- Using `lm()` in R

**Reading:** *Regression and Other Stories*, Chapters 6 & 7.1

---

## Questions? {background-color="#003660"}

::: {style="text-align: center; margin-top: 2em;"}


Come to office Hours tomorrow! 

Tuesday 10 – 11 AM @ 


:::
