---
title: "Lecture 2: Fitting the Simple Linear Model"
subtitle: "PSTAT 126: Linear Regression"
author: "Annie Adams"
institute: "UC Santa Barbara"
date: "9/30/2026"
format:
  revealjs:
    theme: [default, ../styles.scss]
    slide-number: true
    preview-links: auto
    footer: "PSTAT 126 — Fall 2026 · UC Santa Barbara"
    logo: ""
    chalkboard: true
    title-slide-attributes:
      data-background-color: "#003660"
---

## Review: The Simple Linear Model

From Monday, the simple linear regression model is:

$$y_i = \beta_0 + \beta_1 x_i + \varepsilon_i, \quad \varepsilon_i \overset{\text{iid}}\sim N(0, \sigma^2)$$

- $\beta_1$ is expected change in $y$ per one-unit increase in $x$
- $\beta_0$ is expected value of $y$ when $x = 0$
- The fitted line $\widehat{y} = \widehat{\beta}_0 + \widehat{\beta}_1 x$ summarizes the relationship

**Today:** How do we estimate $\widehat{\beta}_0$, $\widehat{\beta}_1$, and $\sigma$ from data?

---

## Learning Objectives 

::: {style="font-size: 31px"}

By the end of today you will be able to:

- Explain the least squares criterion and define the residual sum of squares
- Derive the OLS estimators $\widehat{\beta}_0$ and $\widehat{\beta}_1$ by solving the normal equations
- Compute fitted values and residuals from the estimated line
- Connect $\widehat{\beta}_1$ to the sample correlation between $x$ and $y$
- Show that the OLS estimators are linear combinations of $y_1, \ldots, y_n$
- Show that the OLS estimators are unbiased: $\mathbb{E}(\widehat{\beta}_0) = \beta_0$ and $\mathbb{E}(\widehat{\beta}_1) = \beta_1$
:::

---

## Least Squares Estimation {background-color="#003660"}



## Mini Refresher: Derivatives {.smaller}

**Power rule:** $\frac{d}{dx} x^n = nx^{n-1}$

**Chain rule:** $\frac{d}{dx} f(g(x)) = f'(g(x)) \cdot g'(x)$

**Partial derivatives:** treat all other variables as constants and differentiate with respect to one variable at a time.

**Example:** $\dfrac{\partial}{\partial \beta_0}(y - \beta_0 - \beta_1 x)^2 = 2(y - \beta_0 - \beta_1 x) \cdot (-1)$


## Practice Problems

:::{.fragment}
::: callout-tip
## Practice Problems

:::{.nonincremental}
::: {style="font-size: 30px"}

Find the derivates of the following:

1.$x^6$

2. $y = (x^2 + 1)^3$

Find the partial derivative of the following:

1. $f(\beta_0, \beta_1) = (y - \beta_0 - \beta_1 x)^2$, find $\dfrac{\partial f}{\partial \beta_1}$

```{r}
#| echo: false
countdown::countdown(
  minutes = 3,
  play_sound = TRUE,
  color_border              = "#FEBC11",
  color_text                = "#003660",
  color_running_background  = "#003660",
  color_running_text        = "#FFFFFF",
  color_finished_background = "#FEBC11",
  color_finished_text       = "#003660",
  font_size = "2em",
)
```

:::
:::
:::
:::


---

## Mini Refresher: Summation Rules {.smaller}

**Derivatives through sums:**
$$\frac{\partial}{\partial \beta_0}\sum_{i=1}^n f_i(\beta_0) = \sum_{i=1}^n \frac{\partial}{\partial \beta_0} f_i(\beta_0)$$

**Constants in sums:**

- $\displaystyle\sum_{i=1}^n c = nc$ — a constant summed $n$ times is just $n \times c$
- $\displaystyle\sum_{i=1}^n c \cdot x_i = c\sum_{i=1}^n x_i$ — a constant factor pulls out of the sum

::: {.callout-note}
These are all we need to minimize RSS — apply the chain rule, pass the derivative through the sum, set to zero, and simplify.
:::


## Practice Problems
:::{.fragment}
::: callout-tip
## Practice Problems

:::{.nonincremental}
::: {style="font-size: 30px"}

Simplify the following:

1. $\displaystyle\sum_{i=1}^{5} 4$

2. $\displaystyle\sum_{i=1}^{n} 3x_i$

3. $\dfrac{\partial}{\partial \beta_0}\displaystyle\sum_{i=1}^{n}(y_i - \beta_0)^2$

```{r}
#| echo: false
countdown::countdown(
  minutes = 3,
  play_sound = TRUE,
  color_border              = "#FEBC11",
  color_text                = "#003660",
  color_running_background  = "#003660",
  color_running_text        = "#FFFFFF",
  color_finished_background = "#FEBC11",
  color_finished_text       = "#003660",
  font_size = "2em",
)
```

:::
:::
:::
:::

---

## How do we pick $\beta_0$ and $\beta_1$? {.smaller}

We need a rule to pick the "best" $\widehat{\beta}_0$ and $\widehat{\beta}_1$. To do so, we want to make prediction errors small across all observations simultaneously.

The **residual** for observation $i$ is:
$$e_i = y_i - \widehat{y}_i$$

where $\widehat{y}_i$ is the **predicted value** $\widehat{\beta}_0 + \widehat{\beta}_1 x_i$

To find the best line, we minimize the **Residual Sum of Squares**:
$$\text{RSS} = \sum_{i=1}^n e_i^2 = \sum_{i=1}^n (y_i - \widehat{y}_i)^2$$

We square the residuals so positive and negative errors don't cancel, and so large errors are penalized more heavily.

---

## The Least Squares Criterion {.smaller}


From the previous slide, $e_i = y_i - \widehat{y}_i$. Since the candidate line predicts $\widehat{y}_i = \beta_0 + \beta_1 x_i$, substituting gives:

$$e_i = y_i - \beta_0 - \beta_1 x_i$$

So the **Residual Sum of Squares** is:

$$\text{RSS}(\beta_0, \beta_1) = \sum_{i=1}^{n} e_i^2 = \sum_{i=1}^{n}(y_i - \beta_0 - \beta_1 x_i)^2$$

**Goal:** choose $\widehat{\beta}_0$ and $\widehat{\beta}_1$ to minimize RSS.

---

## The Normal Equations {.smaller}

We have **2 unknowns** ($\beta_0$ and $\beta_1$) and need **2 equations** to pin them down.

Setting $\dfrac{\partial \text{RSS}}{\partial \beta_0} = 0$ and $\dfrac{\partial \text{RSS}}{\partial \beta_1} = 0$ each gives one equation — together they form a **system of two linear equations** called the **normal equations**.

Recall: $\text{RSS}(\beta_0, \beta_1) = \displaystyle\sum_{i=1}^{n}(y_i - \beta_0 - \beta_1 x_i)^2$

**Strategy:** differentiate RSS with respect to each parameter, set each derivative to zero, then use the chain rule and summation rule to simplify.

---

## The Normal Equations: Derivation {.smaller}

:::: {.columns style="font-size: 0.65em;"}

::: {.column width="50%"}

*Set $\partial\text{RSS}/\partial\beta_0 = 0$:*
$$\frac{\partial}{\partial \beta_0}\sum_{i=1}^{n}(y_i - \beta_0 - \beta_1 x_i)^2 = 0$$

*Pass derivative through sum (summation rule):*
<div style="height: 8em;"></div>

*Chain rule:*
<div style="height: 8em;"></div>

*Factor out $-2$, then divide both sides by $-2$:*
<div style="height: 8em;"></div>

:::

::: {.column width="50%"}

*Set $\partial\text{RSS}/\partial\beta_1 = 0$:*
$$\frac{\partial}{\partial \beta_1}\sum_{i=1}^{n}(y_i - \beta_0 - \beta_1 x_i)^2 = 0$$

*Pass derivative through sum (summation rule):*
<div style="height: 8em;"></div>

*Chain rule:*
<div style="height: 8em;"></div>

*Factor out $-2$, then divide both sides by $-2$:*
<div style="height: 8em;"></div>

:::

::::

---

## The Normal Equations: Derivation (Solution) {.smaller}

:::: {.columns style="font-size: 0.65em;"}

::: {.column width="50%"}

*Set $\partial\text{RSS}/\partial\beta_0 = 0$:*
$$\frac{\partial}{\partial \beta_0}\sum_{i=1}^{n}(y_i - \beta_0 - \beta_1 x_i)^2 = 0$$

*Pass derivative through sum (summation rule):*
$$\sum_{i=1}^{n} \frac{\partial}{\partial \beta_0}(y_i - \beta_0 - \beta_1 x_i)^2 = 0$$

*Chain rule:*
$$\sum_{i=1}^{n} 2(y_i - \beta_0 - \beta_1 x_i) \cdot (-1) = 0$$

*Factor out $-2$, then divide both sides by $-2$:*
$$\sum_{i=1}^n (y_i - \beta_0 - \beta_1 x_i) = 0$$

:::

::: {.column width="50%"}

*Set $\partial\text{RSS}/\partial\beta_1 = 0$:*
$$\frac{\partial}{\partial \beta_1}\sum_{i=1}^{n}(y_i - \beta_0 - \beta_1 x_i)^2 = 0$$

*Pass derivative through sum (summation rule):*
$$\sum_{i=1}^{n} \frac{\partial}{\partial \beta_1}(y_i - \beta_0 - \beta_1 x_i)^2 = 0$$

*Chain rule:*
$$\sum_{i=1}^{n} 2(y_i - \beta_0 - \beta_1 x_i) \cdot (-x_i) = 0$$

*Factor out $-2$, then divide both sides by $-2$:*
$$\sum_{i=1}^n x_i(y_i - \beta_0 - \beta_1 x_i) = 0$$

:::

::::

---

## Solving the Normal Equations: Step 1

**Solve equation 1 for $\widehat{\beta}_0$:**

<div style="height: 6em;"></div>

<div style="height: 6em;"></div>

---

## Solving the Normal Equations: Step 1 (Solution)

**Solve equation 1 for $\widehat{\beta}_0$:**

$$\sum_{i=1}^n(y_i - \beta_0 - \beta_1 x_i) = 0 \;\Rightarrow\; n\bar{y} - n\beta_0 - n\beta_1\bar{x} = 0$$

$$\boxed{\widehat{\beta}_0 = \bar{y} - \widehat{\beta}_1\bar{x}}$$

---

## Solving the Normal Equations: Step 2 {.smaller}

::: {style="font-size: 0.82em;"}
**Substitute into equation 2 and solve for $\widehat{\beta}_1$:**

Start with equation 2 and distribute the sum:
<div style="height: 6em;"></div>

Substitute $\beta_0 = \bar{y} - \beta_1\bar{x}$ and use $\sum x_i = n\bar{x}$:
<div style="height: 6em;"></div>

Expand the multiplication:
<div style="height: 6em;"></div>
:::

---

## Solving the Normal Equations: Step 2 (Solution) {.smaller}

::: {style="font-size: 0.82em;"}
**Substitute into equation 2 and solve for $\widehat{\beta}_1$:**

Start with equation 2 and distribute the sum:
$$\sum_{i=1}^n x_i y_i - \beta_0\sum_{i=1}^n x_i - \beta_1\sum_{i=1}^n x_i^2 = 0$$

Substitute $\beta_0 = \bar{y} - \beta_1\bar{x}$ and use $\sum x_i = n\bar{x}$:
$$\sum_{i=1}^n x_i y_i - (\bar{y} - \beta_1\bar{x})\,n\bar{x} - \beta_1\sum_{i=1}^n x_i^2 = 0$$

Expand the multiplication:
$$\sum_{i=1}^n x_i y_i - n\bar{x}\bar{y} + n\bar{x}^2\beta_1 - \beta_1\sum_{i=1}^n x_i^2 = 0$$
:::

---

## Solving the Normal Equations: Step 2 (cont.) {.smaller}

::: {style="font-size: 0.82em;"}
Collect $\beta_1$ terms on the right:
<div style="height: 6em;"></div>

Divide both sides by $\left(\sum x_i^2 - n\bar{x}^2\right)$:
<div style="height: 8em;"></div>
:::

---

## Solving the Normal Equations: Step 2 (cont.) (Solution) {.smaller}

::: {style="font-size: 0.82em;"}
Collect $\beta_1$ terms on the right:
$$\sum_{i=1}^n x_i y_i - n\bar{x}\bar{y} = \beta_1\!\left(\sum_{i=1}^n x_i^2 - n\bar{x}^2\right)$$

Divide both sides by $\left(\sum x_i^2 - n\bar{x}^2\right)$:
$$\boxed{\widehat{\beta}_1 = \frac{\displaystyle\sum_{i=1}^n x_i y_i - n\bar{x}\bar{y}}{\displaystyle\sum_{i=1}^n x_i^2 - n\bar{x}^2} = \frac{S_{xy}}{S_{xx}}}$$
:::

---

## Equivalent Form of $\widehat{\beta}_1$ {.smaller}

The numerator and denominator have a cleaner equivalent form. Expanding $\sum(x_i - \bar{x})(y_i - \bar{y})$:

$$\sum(x_i - \bar{x})(y_i - \bar{y}) = \sum x_iy_i - \bar{y}\sum x_i - \bar{x}\sum y_i + n\bar{x}\bar{y}$$

Since $\sum x_i = n\bar{x}$ and $\sum y_i = n\bar{y}$:

$$= \sum x_iy_i - n\bar{x}\bar{y} - n\bar{x}\bar{y} + n\bar{x}\bar{y} = \sum x_iy_i - n\bar{x}\bar{y}$$

The same identity holds for the denominator: $\displaystyle\sum(x_i - \bar{x})^2 = \sum x_i^2 - n\bar{x}^2$

So the two forms are equivalent. We will use the simpler form on the next slide.

---

## The OLS Estimators {.smaller}

::: {style="font-size: 22px"}

Solving the normal equations yields:

$$\widehat{\beta}_1 = \frac{\displaystyle\sum_{i=1}^n (x_i - \bar{x})(y_i - \bar{y})}{\displaystyle\sum_{i=1}^n (x_i - \bar{x})^2} = \frac{S_{xy}}{S_{xx}}$$

$$\widehat{\beta}_0 = \bar{y} - \widehat{\beta}_1 \bar{x}$$

where:

$$S_{xy} = \sum_{i=1}^n (x_i - \bar{x})(y_i - \bar{y}) \quad \text{(co-variation of } x \text{ and } y \text{ around their means)}$$

$$S_{xx} = \sum_{i=1}^n (x_i - \bar{x})^2 \quad \text{(total spread of } x \text{ around its mean)}$$

:::
---

## Connection to Correlation {.smaller}


The OLS slope has a useful rewrite in terms of correlation:

$$\widehat{\beta}_1 = r \cdot \frac{s_y}{s_x}$$

where $r$ is the sample correlation between $x$ and $y$, and $s_x$, $s_y$ are sample standard deviations.

::: {.incremental}
- The sign of $\widehat{\beta}_1$ always equals the sign of $r$
- When $r = 0$ $\rightarrow$ slope is 0 $\rightarrow$ no linear relationship
- When $|r| = 1$ $\rightarrow$ all points lie exactly on a line $\rightarrow$ perfect fit
:::

---

## Practice: Computing OLS Estimates {.smaller}

A researcher records daily exercise hours ($x$) and wellness score ($y$, out of 15) for $n = 5$ students:

:::: {.columns}

::: {.column width="52%"}

| $i$ | $x_i$ | $y_i$ | $\widehat{y}_i$ | $\hat{e}_i$ |
|:---:|:------:|:------:|:---------------:|:-----------:|
| 1 | 1 | 4 | | |
| 2 | 2 | 3 | | |
| 3 | 3 | 8 | | |
| 4 | 4 | 9 | | |
| 5 | 5 | 11 | | |

:::

::: {.column width="48%"}

**Compute:**

1. $\widehat{\beta}_1 = S_{xy} / S_{xx}$

2. $\widehat{\beta}_0 = \bar{y} - \widehat{\beta}_1\bar{x}$

3. Fill in $\widehat{y}_i$ using $\widehat{y}_i = \widehat{\beta}_0 + \widehat{\beta}_1 x_i$

4. Fill in $\hat{e}_i$ using $\hat{e}_i = y_i - \widehat{y}_i$

```{r}
#| echo: false
countdown::countdown(
  minutes = 7,
  play_sound = TRUE,
  color_border              = "#FEBC11",
  color_text                = "#003660",
  color_running_background  = "#003660",
  color_running_text        = "#FFFFFF",
  color_finished_background = "#FEBC11",
  color_finished_text       = "#003660",
  font_size = "2em",
)
```

:::

::::

---

## Practice: Solution {.smaller}

**Finding $\bar{x}$, $\bar{y}$, $S_{xx}$, $S_{xy}$:**

$$\bar{x} = \frac{1+2+3+4+5}{5} = 3, \qquad \bar{y} = \frac{4+3+8+9+11}{5} = 7$$

$$S_{xx} = (-2)^2+(-1)^2+0^2+1^2+2^2 = 10$$

$$S_{xy} = (-2)(-3)+(-1)(-4)+(0)(1)+(1)(2)+(2)(4) = 20$$

**Estimates:**

$$\widehat{\beta}_1 = \frac{S_{xy}}{S_{xx}} = \frac{20}{10} = 2, \qquad \widehat{\beta}_0 = \bar{y} - \widehat{\beta}_1\bar{x} = 7 - 2(3) = 1$$

Fitted line: $\;\widehat{y} = 1 + 2x$

---

## Practice: Solution (cont.) {.smaller}

**Fitted values and residuals**

| $i$ | $x_i$ | $y_i$ | $\widehat{y}_i$ | $\hat{e}_i$ |
|:---:|:------:|:------:|:---------------:|:-----------:|
| 1 | 1 | 4 | 3 | 1 |
| 2 | 2 | 3 | 5 | −2 |
| 3 | 3 | 8 | 7 | 1 |
| 4 | 4 | 9 | 9 | 0 |
| 5 | 5 | 11 | 11 | 0 |

---

## Properties of the OLS Estimators {.smaller}

::: {style="font-size: 0.82em;"}
**Notation:** $S_{xx} = \displaystyle\sum_{i=1}^n(x_i - \bar{x})^2$ — total spread of the predictor; $\sigma^2$ — true (unknown) error variance; $\mathbb{E}(\cdot)$ — expected value; $\text{Var}(\cdot)$ — variance.
:::

::: {.incremental}
1. **Linear:** $\widehat{\beta}_0$ and $\widehat{\beta}_1$ are linear functions of $y_1, \ldots, y_n$

2. **Unbiased:** $\mathbb{E}(\widehat{\beta}_0) = \beta_0$ and $\mathbb{E}(\widehat{\beta}_1) = \beta_1$ — on average, the estimates recover the true parameters

3. **Known variances:**
$$\text{Var}(\widehat{\beta}_1) = \frac{\sigma^2}{S_{xx}}, \qquad \text{Var}(\widehat{\beta}_0) = \sigma^2\!\left(\frac{1}{n} + \frac{\bar{x}^2}{S_{xx}}\right)$$

4. **Gauss-Markov (BLUE):** OLS has the smallest variance among all linear unbiased estimators — no other such estimator can do better.
:::

---


## Summary {.smaller}

::: {.incremental}
- **Least squares:** OLS picks $\widehat{\beta}_0$ and $\widehat{\beta}_1$ to minimize $\text{RSS} = \sum(y_i - \widehat{y}_i)^2$
- **Normal equations** give closed-form solutions: $\widehat{\beta}_1 = S_{xy}/S_{xx}$, $\;\widehat{\beta}_0 = \bar{y} - \widehat{\beta}_1\bar{x}$
- **Residuals** $e_i = y_i - \widehat{y}_i$ measure how far each observation falls from the fitted line
:::

---


## Questions? {background-color="#003660"}

::: {style="text-align: center; margin-top: 2em;"}
**Annie Adams**
[aradams@ucsb.edu](mailto:aradams@ucsb.edu)

Office Hours: Tuesday 10 – 11 AM on [Zoom](https://ucsb.zoom.us/j/81194403106?pwd=sQo3HO4iaAnzcY0apRmIyAdC1KbVfT.1)


[Course Website](https://pstat-126.github.io/)
:::
