---
title: "Getting started with R Markdown"
author: "PSTAT126"
date: "Lab 1"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: sentence
---

### Install and load packages

```{r setup, include = FALSE}
# To clean the environment
rm(list=ls())
# default code chunk options
knitr::opts_chunk$set(echo = F,
                      results = 'markup',
                      message = F, 
                      warning = F) 

# Install Packages
# install.packages("faraway")
# install.packages("tidyverse")


# load packages
library(faraway)
library(tidyverse)


```

### Objectives

This short getting started guide introduces the basics of R Markdown (.Rmd) files, which 'knit' together text and code, and some style conventions.

-   Elements of .Rmd: code chunks and text
-   Executing codes within .Rmd files in RStudio
-   Code style guidelines
-   Controlling code chunk display in knitted documents
-   Controlling text display in knitted documents
-   Math display
-   Report formatting and code appendices

# R Markdown: integrating text and code

R Markdown allows for easy building of documents that integrate codes and code results with text.
R Markdown files (.Rmd) contain code chunks and regular text.
The code chunks are executable from within RStudio as the document is being composed and edited.
This makes .Rmd a great choice for projects involving data analysis in R, because one can simultaneously compose codes *and* narrative without having to keep track of multiple files or documents.
This is very conducive to reproducible research and version control.

When an R Markdown file is 'knitted' in RStudio, a report is generated that displays text and code and code results according to author-specified parameters in the document.
Your first task in this lab is to experiment with code chunk options and text markdown in the .Rmd file that control the display of codes, text, and results in the knitted document.

## Codes: the "R" in R Markdown

First we'll examine aspects of the .Rmd file that pertain to R codes, output, and graphics.

## Basic functions

While composing a document, code chunks can be written and executed from within the .Rmd file in RStudio.
Below this paragraph in the .Rmd file is an example code chunk with some simple commands.
Try running individual lines from the .Rmd file for this document within RStudio.
The results will appear below the chunk and in the console.

*Reminder*: you can execute code by placing your cursor on the line you wish to run and pressing `CTRL + ENTER` or `CMD + ENTER`.

*Tip*: If you wish to run an entire chunk, use the 'play' icon in green that appears in the upper right corner of the chunk.

```{r example-chunk, results ='hide'}
# concatenation `c()`
c(1, 3, 2, 5)

# assignment `<-`
x <- c(1, 3, 2, 5)
x

# sequence of numbers starting from 1 until 5 by 1
x <- seq(from = 1, to = 5, by = 1) 

# repeats 1 four times
x <- rep(1, times=4) 

# repeats each value in 1,2,3,4 two times
x <- rep(1:4, each=2) 

# check object structure `str()`
str(x)
x[1]

mean(x)

sd(x)

var(x)

sort(x)

summary(x)

# vectorized arithmetic and logic
y <- x + 1
y
x > y


```

Now notice that the code chunk is set off by three back-ticks and some text in braces \`\`\``{some text}`.
The first part of the text in braces is *always* `r`; optionally, a name for the chunk (above, `example-chunk`); then a comma, followed by **code chunk options** that control how the code will be displayed in the knitted document.

## Knitting document

Click the 'Knit' button with the ball-of-yarn icon and have a look at the knitted document.
For the code chunk above, nothing appears!
To change this behavior we can change the code chunk options.
To get started, let's tinker with those options.

To knit to pdf or HTML, go all the way to the top of the document, and change the `output` part of the header block to `html_document` or to `pdf_document` respectively.

### R code chunks options

-   **echo:** Display code in output document (default = TRUE)
-   **eval:** Run code in chunk (default = TRUE)
-   **message:** Display code messages in document (default = TRUE)
-   **warning:** Display code warnings in document (default = TRUE)
-   **fig.align:** "left", "right", or "center".
-   **fig.height**, **fig.width**: Dimensions of plots in inches.

### Your turn

Make each of the following changes one by one, and re-knit after each change to examine the effect on the knitted document.

-   Change the name of the chunk to your favorite author's last name. After knitting, examine the 'Render' tab in RStudio and look for where the new name appears. Try introducing a command in the code chunk that will produce an error, and knit again. How does naming the chunk prove to helpful in this instance?

**Answer:**

-   Change the `results = 'hide'` option to `results = 'markup'`. Also try `results = 'asis'`. What is the difference?

**Answer:**

-   Remove the `results` option altogether.What is the default behavior? Now go all the way to the top of the .Rmd file and examine the `setup` chunk. Try changing the `results` option there to one of the other options. After examining the effect on the knitted document, revert to `results = 'markup'`.

**Answer:**

-   Try adding `echo = T` to the example code chunk. What happens?

**Answer:**

-   Configure the code chunk options so that the knitted document shows the codes themselves but not the results.

## Matrices

```{r, echo=TRUE}
# using matrix() function
mat <- matrix(c(3,2,5,3,1,4,7,4,9), nrow = 3)
mat

# fill matrix by row
mat <- matrix(c(3,2,5,3,1,4,7,4,9), byrow=T, nrow = 3) 
mat

# Using rbind() or cbind() Function
x <- 1:5
y <- 2:6
mat <- cbind(x,y) # create a matrix by combining vectors column-wise
mat

mat <- rbind(x,y) # create a matrix by combining vectors row-wise
mat

# How to name the rows and columns of an R matrix?

# Using dimnames argument in matrix() function
rnames <- c("row1","row2","row3")
cnames <- c("col1","col2","col3")
mat <- matrix(c(3,2,5,3,1,4,7,4,9), nrow=3, dimnames=list(rnames,cnames))
mat

# Using rownames() or colnames() functions
mat <- matrix(c(3,2,5,3,1,4,7,4,9), nrow = 3)
rownames(mat) <- c("row1","row2","row3")
colnames(mat) <- c("col1","col2","col3")
mat

mat%*%mat # matrix multiplication

solve(mat) # Matrix inversion for non-singular matrices

diag(mat) # extract the diagonal elements of the matrix

t(mat) # transpose of matrix
```

## Working with DataFrame

```{r, echo=TRUE}
subject <- c("English", "Maths", "Chemistry", "Physics")
percentage <- c(80, 100, 85, 95)
df <- data.frame(subject, percentage)
df

# Appending a column to a DataFrame
df$GPA <-c(3.0, 4.0, 3.0, 4.0)
df
```

### Plots

There are multiple ways to plot in R, either using base R or `ggplot2`

```{r, echo=TRUE}
head(eco)
#?eco

plot(eco$usborn, eco$income,
xlab = "Percentage of Babies Born in US", ylab = "Income ($)",
main = "Percentage of US Born per Income")
```

```{r lecture-plot, fig.show ='hide', fig.align='center', fig.width=7, fig.height=5}
library(ggplot2)

# scatterplot
ggplot(data = eco, mapping = aes(x = usborn, y = income)) +
  geom_point() +
  labs(title = "Percentage of US Born per Income",
       y = "Income ($)",
       x = "Percentage of Babies Born in US") +
  theme()
#?labs

```

Now once the plot looks nice, try knitting.
Notice that the plot *does not* appear in the knitted document.
This is due to the `fig.show = 'hide'` option; remove this and re-knit, and the plot should appear.

#### Your turn

Often it is necessary to resize plots for knitting to avoid oversize or cut-off images.
Try the following chunk options one by one and knit between each change to take note of the effect in the knitted document.

i.  Add the option `fig.width = 2` and examine the output. Tinker with the option by changing the number until you find a display width that you like.
ii. Add the option `fig.height = ...`, specifying a number of your choosing for the height. Play until you find a display height that you like.
iii. Notice that the alignment is off-center. Try adding `fig.align = 'center'`.

There are many other code chunk options, but those come up particularly often.
The above and other options are summarized in the [R Markdown cheatsheet](https://rstudio.com/resources/cheatsheets/).

### Code style guidelines

A few points of style can dramatically improve readability:

-   `<-` is used for assignment (not `=`, used to specify function arguments)
-   every comma is followed by a space
-   spaces on either side of `=` and `<-`, logical operations, and arithmetic operators ('infix' operators)
-   short explanatory comments above groups of lines preceded by a blank line
-   use descriptive names with `_` to introduce spaces separating name components
-   break long commands over multiple lines
-   specify function arguments explicitly by name

## RStudio Settings Tips

A few RStudio settings that make writing and reading code much easier.

### Rainbow Parentheses

Rainbow parentheses color-code matching pairs of parentheses, brackets, and braces, making it much easier to spot mismatched or missing delimiters.

To enable: go to **Tools \> Global Options \> Code \> Display** and check the box for **Rainbow parentheses**.

Once enabled, nested parentheses like `mean(c(1, sd(x), var(y)))` will each appear in a distinct color, so you can immediately see which opening and closing symbols are paired.

### Render Output in Viewer Pane

By default, knitting an .Rmd file opens the output in a new window or browser tab.
You can instead have it appear in the **Viewer** pane inside RStudio, which keeps everything in one place.

To enable: go to **Tools \> Global Options \> R Markdown** and set **Show output preview in** to **Viewer Pane**.

## Brief Tidyverse intro

Other useful functions in the tidyverse are `group_by()`,`arrange()`,`mutate()` , and `summarise()` For more info on tidyverse function go to <https://r4ds.had.co.nz/transform.html/>.
You can use `glimpse()` (from tidyverse package) or `str()` to get a quick look at a dataset.

```{r, echo=TRUE}

# Install Tidyverse and dplyr if you haven't already
#install.packages("tidyverse")
#install.packages("dplyr")


library(tidyverse)
library(dplyr)

eco %>%
dplyr::select(usborn) %>% # select usborn column
filter(usborn <= 0.9) %>% # filter usborn leq 0.9
slice(1:5) # print out only first 5 rows
```

The $\%>\%$ is commonly referred to as the "pipe operator" and is used to make code more readable by chaining together a sequence of operations.

```{r, echo=TRUE}
eco %>%
filter(usborn <= 0.9) %>% # filter usborn leq 0.9
ggplot() +
# scatterplot with usborn on x axis and income on y
geom_point(aes(x = usborn, y = income),
# orange color data points. alpha changes the transparency of the points
color = "orange", alpha = 0.7)
```

## Distributions

Can obtain random sample from a specific distribution: Type in ?d
istribution in console to see all the different distributions in R.

-   p for 'probability', the cumulative distribution function (c. d. f.)
-   q for 'quantile', the inverse c. d. f.
-   d for 'density', the density function (p. f. or p. d. f.)
-   r for 'random', a random variable having the specified distribution

```{r, echo=TRUE}
# Type  '?distribution' in console to see all the different distributions in R.
# Remember to set seed
set.seed(555)
# r for “random”, a random variable having the specified distribution
rnorm(10, mean = 0, sd = 1)
runif(10, min = 0, max = 1)

normal_sample <- rnorm(1000, mean = 0, sd = 1)
mean(normal_sample) # mean of sample
hist(normal_sample) #

ggplot() +
geom_histogram(aes(x = normal_sample), # using ggplot to make histograms
bins = 10, color = "white", # control the number of bins
fill = "darkgreen")

```

## Text: the 'Markdown' in R Markdown

Now we'll look over the text aspects of .Rmd documents.

### Narrative text

You've probably noticed by now that headers are set off by `#`, `##`, `###` in the .Rmd document, and these are rendered in nice format in the knitted document.
These are examples of text markdown elements, which are used to control the display of text in the knitted document.
Some common ones are:

-   `#` header, number of hashes specifies header level
-   `*italic text*` renders in *italic*
-   `**bold text**` renders in **bold**
-   `*, -, +` set off itemize list elements
    -   sub-item
-   `1., 2., ..` set off numbered lists
-   `[link text](url)` creates hyperlinks

These and other options are summarized in the [R Markdown cheatsheet](https://rstudio.com/resources/cheatsheets/).

### Mathematical notation

R Markdown documents also integrate math display using \LaTeX math expressions.
For inline math environments, enclose expressions in `$`, such as `$\left(\frac{1}{x}\right)$`, which appears as $\left(\frac{1}{x}\right)$.
The .Rmd file has dynamic previewing while building a document; hovering your cursor over the expression while editing will show you what it will look like in the generated document.

Equation blocks can be created by enclosing math expressions in `$$`.
For example, `$$Y = \mathbf{x}\beta + \epsilon$$` displays as: $$Y = \mathbf{X}\beta + \epsilon$$

For a list of LaTeX math syntax, see [wikibooks](https://en.wikibooks.org/wiki/LaTeX/Mathematics).
(These expressions also work on Nectir using the same syntax!)

#### Your turn

Write the density of a Gaussian random variable in an equation block.
Below that, write the mean and variance in inline math mode (preferably in a short sentence).

$$
f(x) = \frac{1}{\sqrt{2 \pi \sigma^2}} e^{\frac{-(x-\mu)^2}{2\sigma^2}}
$$

The mean is $\mathbb{E}[X] = \mu$ and the variance is $\text{Var}(X) = \sigma^2$.

# Submission Instructions

Before submitting, knit your completed lab to PDF:

1.  At the top of this file, make sure the `output` field in the YAML header includes `pdf_document: default`.
2.  Click the **Knit** button and select **Knit to PDF**.
3.  Before knitting, double-check that all code chunk options (`echo`, `results`, `fig.show`, `fig.width`, `fig.height`, etc.) are set exactly as the lab instructions asked — do not change them just to clean up the output.
4.  Once the PDF has been generated, review it to make sure all requested output, plots, and written answers appear correctly.
5.  Upload the PDF to **Gradescope** under the assignment for this lab.

# Extra: code appendix

Imagine you're writing a report of some data analysis (*e.g.*, a homework for this class).
While it is lovely that R Markdown allows you to integrate your analysis and results directly into your written report, very likely there are many code chunks that need to be run but not necessarily shown in your report to answer the questions of interest.

It is good practice to suppress codes or output that your readers don't need to see (with the `results = 'hide'` and/or `include = F` chunk options) in the body of your report.
Examples include:

-   setup commands like package loads
-   data import
-   printed data frames

Consider setting the global options to suppress all code and output, so that you have to specifically choose to show a code chunk or result.
This is a better policy than defaulting to showing all code and output, which typically results in an unreadable document.

However, your TA might need to check some of those codes if they spot an error somewhere.
In that case, including an appendix of all codes at the end can be helpful.

```{r appendix, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```

# Submission Instructions

Before submitting, knit your completed lab to PDF:

1. At the top of this file, make sure the `output` field in the YAML header includes `pdf_document: default`.
2. Click the **Knit** button and select **Knit to PDF**.
3. Before knitting, double-check that all code chunk options (`echo`, `results`, `fig.show`, `fig.width`, `fig.height`, etc.) are set exactly as the lab instructions asked — do not change them just to clean up the output.
4. Once the PDF has been generated, review it to make sure all requested output, plots, and written answers appear correctly.
5. Upload the PDF to **Gradescope** under the assignment for this lab.
